---
title: "Entire USA--Artesian Comparison"
author: "Annette Hilton"
date: "05/27/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the continental USA as compared to present day (2010-2021).

The purpose of this Rmd is to: 

1) Determine well-aquifer classification 
2) Compare artesian conditions 
  

## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)
library(knitr)

# Disable scientific notation 

options(scipen=999)
```

# Depth to Confined Investigations 

1. Spatial join modern data (`modern_final.txt` from `modern_data_cleaning.Rmd`) with aquifer file and depth data: processsed in ArcPro (file name: ) 
2. Load into R for analysis 
3. Analyze and determine artesian vs non artesian 
4. Map in ArcPro 

```{r}
# Step 2. 
# Modern data 

modern_depths <- readr::read_csv(here::here("data", "depths", "modern_depth_joined.txt"))

# 1900s data 
early1900s_depths <- readr::read_csv(here::here("data", "depths", "early1900s_depths.txt"))

```

```{r}
# How many wells do we have for each aquifer system that is greater than the confined depth cut off? 

modern_analysis_depth <- modern_depths %>% 
  mutate(depth_m = well_depth_va * 0.3048) %>% 
  mutate(confined_depth = ifelse(depth_m >= DepthConfi, "confined depth", NA))

modern_depth_final <- modern_analysis_depth %>% 
  filter(!is.na(confined_depth)) %>% 
  group_by(Aquifer_1) %>% 
  filter(n() >= 4)

write_tsv(modern_depth_final, here::here("data", "depths", "modern_depth_final.txt"))
```

```{r}
# How many wells do we have for each aquifer system that is greater than the confined depth cut off? 

historical_analysis_depth <- early1900s_depths %>% 
  mutate(depth_m = well_depth * 0.3048) %>% 
  mutate(confined_depth = ifelse(depth_m >= DepthConfi, "confined depth", NA))

historical_depth_final <- historical_analysis_depth %>% 
  filter(!is.na(confined_depth)) %>% 
  group_by(Aquifer_1) %>% 
  filter(n() >= 4)

write_tsv(historical_depth_final, here::here("data", "depths", "historical_depth_final.txt"))
```

How many aquifers? 

```{r}
# Modern data 
aquifer_modern <- modern_depth_final %>% 
  distinct(Aquifer_1, .keep_all = T)

# 1900 data 
aquifer_historical <- historical_depth_final %>% 
  distinct(Aquifer_1, .keep_all = T)

```

Join modern and historical data 

```{r}
both_hist_modern <- aquifer_modern %>% 
  inner_join(aquifer_historical, by = "Aquifer_1")

write_tsv(both_hist_modern, here::here("data", "depths", "both_aquiferstimeperiod.txt"))
```

#### Artesian Calculations 

Only use data where 1900s and modern overlap in aquifer systems (n = 63) 

```{r}
# Make vector of shared aquifer systems 

shared_aquifers <- both_hist_modern %>% 
  pull(Aquifer_1)

# Use the resulting vector to filter target sites from modern and 1900 data frames 

modern_constrained <- modern_depth_final %>%
  filter(Aquifer_1 %in% c(shared_aquifers))

historical_constrained <- historical_depth_final %>%
  filter(Aquifer_1 %in% c(shared_aquifers))
```

Write files to tsv 

```{r}
# Modern 
write_tsv(modern_constrained, here::here("output_data", "depths", "modern_usa_confined.txt"))

# 1900s
write_tsv(historical_constrained, here::here("output_data", "depths", "historical_usa_confined.txt"))
```

### Summary tables 

##### Modern Data 

```{r}
# Modern data

# Confined/unconfined and artesian by aquifer unit

summary_mod_final_art <- modern_usa %>% 
  group_by(aquifer, status, artesian) %>% 
  count(aquifer) %>% 
  filter(!is.na(status))

# Pivot wider 

mod_final_wide <- pivot_wider(summary_mod_final_art, 
                          names_from = c(status, artesian), 
                          values_from = n) %>% 
  mutate(total_n = rowSums(across(where(is.numeric)), na.rm = T)) %>% 
  rowwise() %>% 
  mutate(total_c = sum(c(confined_0, confined_1), na.rm = T)) %>% 
  mutate(per_c = (total_c / total_n) *100) %>% 
  mutate(per_art = (confined_1 / total_c) *100)

# Tidy the table 

mod_final_table <- mod_final_wide %>% 
  select(aquifer, total_n, total_c, per_c, confined_1, per_art, confined_0) %>% 
  replace(is.na(.), 0) %>% 
  mutate_if(is.numeric, round, digits = 2) %>% 
  unite(confined_wells, c("total_c", "per_c"), sep = "  |  ") %>% 
  unite(artesian_wells, c("confined_1", "per_art"), sep = "  |  ") %>% 
  rename("Aquifer Units" = "aquifer", 
         "N wells" = "total_n", 
         "Number of wells defined as confined" = "confined_wells", 
         "Artesian confined flowing wells" = "artesian_wells",
         "Not artesian" = "confined_0") 


# Write to file 
write_csv(mod_final_table, here::here("output_data", "columbia", "mod_final_table.txt"))
  
```

```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(mod_final_table, caption = "Table 1. Modern (2010-2021) confined and artesian wells of the Columbia Plateau")
```

##### 1900s Data 

```{r}
# 1900s data

# Confined/unconfined by aquifer unit 
summary_1900_final <- confined_criteria_1900 %>% 
  group_by(aquifer, status, artesian) %>% 
  count(aquifer) 

# Pivot wider 

early1900s_final_wide <- pivot_wider(summary_1900_final, 
                          names_from = c(status, artesian), 
                          values_from = n) %>% 
  mutate(total_n = rowSums(across(where(is.numeric)), na.rm = T)) %>% 
  rowwise() %>% 
  mutate(total_c = sum(c(confined_0, confined_1), na.rm = T)) %>% 
  mutate(per_c = (total_c / total_n) *100) %>% 
  mutate(per_art = (confined_1 / total_c) *100)

# Tidy the table 

early1900s_final_table <- early1900s_final_wide %>% 
  select(aquifer, total_n, total_c, per_c, confined_1, per_art, confined_0) %>% 
  filter(!is.na(aquifer)) %>% 
  replace(is.na(.), 0) %>% 
  mutate_if(is.numeric, round, digits = 2) %>% 
  unite(confined_wells, c("total_c", "per_c"), sep = "  |  ") %>% 
  unite(artesian_wells, c("confined_1", "per_art"), sep = "  |  ") %>% 
  rename("Aquifer Units" = "aquifer", 
         "N wells" = "total_n", 
         "Number of wells defined as confined" = "confined_wells", 
         "Artesian confined flowing wells" = "artesian_wells",
         "Not artesian" = "confined_0") 

# Write to file 
write_csv(early1900s_final_table, here::here("output_data", "columbia", "early1900s_final_table.txt"))

```


```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(early1900s_final_table, caption = "Table 2. Early 1900s confined and artesian wells of the Mississippi Embayment")
```

### Visualization 

Confined and unconfined well depth 

```{r}
# Data frame for plots 
ggplot_modern <- cp_modern_check %>% 
  filter(!check %in% c("wrong_c", "wrong_uc")) %>% 
  filter(!is.na(check))

# Violin plot comparing confined, unconfined and USGS data/no data 
ggplot(ggplot_modern, 
       aes(x = status, 
           y = well_depth_va, 
           fill = check)) + 
  geom_violin()

```
