---
title: "Early Well Records (~1880-1910) Data Cleaning"
author: "Annette Hilton"
date: "11/12/2021"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Continental USA early 1900s Data & Analysis 

## Attach Packages

```{r}
# Attach packages 

library(tidyverse)
library(here)
library(janitor)
library(dplyr)


# Disable scientific notation 

options(scipen=999)
```

# Read in Data

## Early Well Data Sources 

1) Early 1900s CV wells data: manually compiled by Jasechko lab, 2021 (June-October) 
  - Darton, 1905
  - Fuller, 1905
  - Fuller, 1906
  - Crider & Johnson, 1906
  - Veatch, 1906 
  
2) Mendenhall et al., 1916 (https://www.sciencebase.gov/catalog/item/58923e21e4b072a7ac143c11)

3) USGS Groundwater Wells (See Rproject 'usa_groundwater_data' by Hilton for more details)

### 1--Early data from Jasechko lab 

```{r, message = FALSE}
# Specify NA values 

na_values <- c("na", "Na", "NA")

# early 1900s data 
# specify each column type 
# specify how to deal with NAs 

jl_early1900s_raw <- readr::read_tsv(here::here("data", "1900s", "1900s_wells.txt"), 
                                     col_types = list(
                                       id = col_character(),
                                       state = col_character(), 
                                       county = col_character(), 
                                       city = col_character(), 
                                       specific_location = col_character(), 
                                       township = col_character(), 
                                       range = col_character(), 
                                       section = col_character(), 
                                       Latitude = col_double(), 
                                       Longitude = col_double(), 
                                       min_depth_ft = col_double(), 
                                       max_depth_ft = col_double(), 
                                       depth_sign = col_character(), 
                                       depth_text = col_character(), 
                                       min_height_water_ft = col_double(), 
                                       max_height_water_ft = col_double(), 
                                       height_water_text = col_character(), 
                                       remarks = col_character(), 
                                       citation = col_character()
                                     ), 
                                     na = na_values)

# early 1900s data with lat/long offset for appropriate wells (only those wells) 

latlong_adjust <- readr::read_tsv(here::here("data", "1900s", "1900s_latlong_offset.txt"))
latlong_adjust$id <- as.character(latlong_adjust$id)
```

### 2-- Mendenhall et al., 1916 

```{r}
mendenhall_raw <- readr::read_csv(here::here("data", "1900s", "MENDENHALL_DATA (1).csv")) %>% 
  clean_names()
```

### 3---USGS Groundwater Wells 


```{r}
# Read in the full dataset for all well water measurements for the USA 

entire_usa_gw <- readr::read_tsv(here::here("data", "entire_usa_2022.txt")) 

```


# Tidy Data 

- Lowercase/clean names
- Remove extra blank columns 
- Remove values without well depth value/depth is zero  
- Remove values without a lat/long 
- Rename columns to be consistent with other data frames 
- Add artesian conditions/columns 


### 1-- Early 1900s Jasechko Lab 

```{r}
# Join two dataframes to incorporate adjusted lat/long 

joined_jl <- jl_early1900s_raw %>% 
  full_join(latlong_adjust, by = "id")

# Tidy data frame 

early_1900s_tidy <- joined_jl %>% 
  clean_names() %>% 
  select(-x20:-x46, -citation_y, -state_y) %>% 
  filter(!is.na(max_depth_ft),
         !is.na(latitude)) %>% 
  filter(!(is.na(max_height_water_ft) & is.na(height_water_text)), 
         !height_water_text %in% c("+-", "+", "-")) %>% 
  rename(well_depth_ft = max_depth_ft, 
         water_level_ft = max_height_water_ft) %>% 
  mutate(artesian = ifelse(water_level_ft >= 0 | !is.na(height_water_text), 1, 0)) %>% 
  mutate(latitude_adj = ifelse(!is.na(new_lat), new_lat, latitude), 
         longitude_adj = ifelse(!is.na(new_lon), new_lon, longitude)) %>% 
  select(-latitude, -longitude, -new_lat:-diff_lon) %>% 
  rename(latitude = latitude_adj, 
         longitude = longitude_adj) %>% 
  filter(!is.na(well_depth_ft))

early_1900s_tidy$id<- sub("^", "w", early_1900s_tidy$id)


```

### 2-- Mendenhall et al., 1916

```{r}
# mendenhall_tidy <- mendenhall_raw %>% 
#   rename(id = site_id, 
#          latitude = lat_dd, 
#          longitude = long_dd, 
#          water_level_ft = water_depth_ft) %>% 
#   mutate(correct_water_levelft = water_depth_m*3.28084, 
#            artesian = ifelse(correct_water_levelft == 0 | water_level_ft == 0, 1, 0)) # WRONG SAD TIMES 
# 
# mendenhall_tidy$id <- sub("^", "w", mendenhall_tidy$id)
# 
# # Quick check for methods (6/23/22) 
# 
# check_men <- mendenhall_tidy %>% 
#   filter(!is.na(water_level_ft)) %>% 
#   filter(!is.na(well_depth_ft)) %>% 
#   filter(!is.na(artesian))
# 
# pull_menden <- check_men %>% 
#   pull(id) 
# 
# check_menden <- check_men %>% 
#   select(id)
# 
# write_tsv(check_menden, here::here("output_data", "cv", "checking_menden.txt"))

```


### 2-- Mendenhall et al., 1916 -- CORRECT, OR HOPEFULLY 

```{r}

mendenhall_tidy <- mendenhall_raw %>% 
  rename(id = real_site_id, 
         latitude = lat_dd, 
         longitude = long_dd, 
         water_level_ft = water_depth_ft) %>% 
  mutate(correct_water_levelft = water_depth_m*3.28084) %>% 
  filter(!(is.na(water_level_ft) & is.na(correct_water_levelft))) %>% 
  filter(!is.na(well_depth_ft)) %>% 
  mutate(final_waterlev = ifelse(is.na(water_level_ft), correct_water_levelft, water_level_ft)) %>%
  mutate(artesian = ifelse(final_waterlev <= 0, 1, 0))

 mendenhall_tidy$id <- sub("^", "w", mendenhall_tidy$id)
```

### 3-- USGS Groundwater Wells 

(USGS email correspondence indicate: neg = artesian, pos = below land surface)

```{r}
# Entire USGS groundwater well data until fall 2019 (last updated) 

usa_early <- entire_usa_gw %>% 
  filter(!well_depth_va == "NA" | !hole_depth_va == "NA") %>% 
  filter(!well_depth_va == 0.00) %>% 
  filter(!lev_dt == "NA") %>% 
  filter(!level_year >1906) %>% 
  filter(!is.na(lev_va))

# Find median water level for wells that have more than one measurement per year 

early_usgs_med <- usa_early %>% 
  group_by(site_no) %>% 
  mutate(median = median(lev_va)) %>% 
  distinct(site_no, .keep_all = TRUE) %>% 
  rename(id = site_no, 
         latitude = dec_lat_va, 
         longitude = dec_long_va, 
         well_depth_ft = well_depth_va, 
         water_level_ft = median) %>% 
  mutate(artesian = ifelse(water_level_ft <= 0, 1, 0)) %>% 
  filter(!is.na(well_depth_ft))

sum_early1 <- early_usgs_med %>% 
  group_by(artesian) %>% 
  count(artesian)

early_usgs_med$id <- sub("^", "w", early_usgs_med$id)
  
```

# Prepare for Analysis 

1. Join data frames (id, lat, long, well depth, water level) 
2. Write for ArcGIS DEM processing 
3. (Separately) process in ArcGIS for DEM values 
4. Import DEM file, merge with original 1900s dataframe 
5. Write file for analyses in `well_classify.Rmd` 


1. Join data frames (id, lat, long, well depth, water level) 
```{r}
# Join data frames 
# First tidy for join 

join_jl <- early_1900s_tidy %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)

join_mendenhall <- mendenhall_tidy %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)

# write for arcgis dem processing (late addition with corrections--6/23/22) 

write_tsv(join_mendenhall, here::here("output_data", "cv", "menden_forarc.txt"))

join_usgs <- early_usgs_med %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)
```

```{r}
# Join data frames 

early1900s_all <- rbind(join_jl, join_usgs)
```

```{r}
# Tidy data
# Remove data without artesian indication 

early1900s_arc <- early1900s_all %>% 
  filter(!is.na(artesian))
  
```

2. Write for ArcGIS DEM processing 

```{r}
# Write now-cleaned data as text file for ArcGIS

write_tsv(early1900s_arc, here::here("output_data", "early1900s_arc.txt"))
```

3. (Separately) process in ArcGIS for DEM values

- On external hard drive, "look busy" 
- `me project` 
- `cv_mendenhall` 
*Extract multi values to points 

4. Import DEM file, merge with original 1900s dataframe 

```{r}
# DEM early 1900s 

early1900s_dem <- readr::read_csv(here::here("data", "dem", "early1900s_dem_wgs84_final.txt"))

# Extra Mendenhall CV data 

cv_dem_mend <- readr::read_csv(here::here("data", "dem", "cv_mend_dem062322.txt"))

# Tidy DEM file 
# Keep only the ID and the DEM value 

early1900s_dem_tidy <- early1900s_dem %>% 
  select(id, ned10m_bilinear_cusa_albers102003) 

cv_dem_mend_tidy <- cv_dem_mend %>% 
  select(id, cali)

# Join with metadata file/full 1900s file 

early1900s_final <- early1900s_dem_tidy %>% 
  inner_join(early1900s_arc, by = "id") 

# Join cv mendenhall files and dem 

help_please <- cv_dem_mend_tidy %>% 
  inner_join(join_mendenhall, by = "id")
```


5. Write file for analysis in `well_classify.Rmd` 

```{r}
# Write final 1900s file, includes DEM and original data   

write_tsv(early1900s_final, here::here("data", "early1900s_final.txt"))
```

# Depth to Confined Investigations (05/18/22) 

1. Spatial join modern data with aquifer file and depth data: processsed in ArcPro (file name: ) 
2. Load into R for analysis 

```{r}
# Step 2. 
early1900s_depths <- readr::read_csv(here::here("data", "depths", "early1900s_depths.txt"))
```

```{r}
# How many wells do we have for each aquifer system that is greater than the confined depth cut off? 

historical_analysis_depth <- early1900s_depths %>% 
  mutate(depth_m = well_depth * 0.3048) %>% 
  mutate(confined_depth = ifelse(depth_m >= DepthConfi, "confined depth", NA))

historical_depth_final <- historical_analysis_depth %>% 
  filter(!is.na(confined_depth)) %>% 
  group_by(Aquifer_1) %>% 
  filter(n() >= 4)

write_tsv(historical_depth_final, here::here("data", "depths", "historical_depth_final.txt"))
```

How many aquifers? 

```{r}
aquifer_historical <- historical_depth_final %>% 
  distinct(Aquifer_1, .keep_all = T)
```

