---
title: "Early Well Records (~1880-1910) Data Cleaning"
author: "Annette Hilton"
date: "11/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Continental USA early 1900s Data & Analysis 

## Attach Packages

```{r}
# Attach packages 

library(parzer)
library(Rcpp)
library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(broom)
library(tibble)
library(dplyr)
library(tidyr)
library(ggpubr)
library(data.table)


# Disable scientific notation 

options(scipen=999)
```

# Read in Data

## Early Well Data Sources 

1) Early 1900s CV wells data: manually compiled by Jasechko lab, 2021 (June-October) 
  - Darton, 1905
  - Fuller, 1905
  - Fuller, 1906
  - Crider & Johnson, 1906
  - Veatch, 1906 
  
2) Mendenhall et al., 1916 (https://www.sciencebase.gov/catalog/item/58923e21e4b072a7ac143c11)

3) USGS Groundwater Wells (See Rproject 'usa_groundwater_data' by Hilton for more details)

### 1--Early data from Jasechko lab 

```{r}
# Specify NA values 

na_values <- c("na", "Na", "NA")

# early 1900s data 
# specify each column type 
# specify how to deal with NAs 

jl_early1900s_raw <- readr::read_tsv(here::here("data", "1900s", "1900s_wells.txt"), 
                                     col_types = list(
                                       id = col_character(),
                                       state = col_character(), 
                                       county = col_character(), 
                                       city = col_character(), 
                                       specific_location = col_character(), 
                                       township = col_character(), 
                                       range = col_character(), 
                                       section = col_character(), 
                                       Latitude = col_double(), 
                                       Longitude = col_double(), 
                                       min_depth_ft = col_double(), 
                                       max_depth_ft = col_double(), 
                                       depth_sign = col_character(), 
                                       depth_text = col_character(), 
                                       min_height_water_ft = col_double(), 
                                       max_height_water_ft = col_double(), 
                                       height_water_text = col_character(), 
                                       remarks = col_character(), 
                                       citation = col_character()
                                     ), 
                                     na = na_values)

# early 1900s data with lat/long offset for appropriate wells (only those wells) 

latlong_adjust <- readr::read_tsv(here::here("data", "1900s", "1900s_latlong_offset.txt"))
latlong_adjust$id <- as.character(latlong_adjust$id)
```

### 2-- Mendenhall et al., 1916 

```{r}
mendenhall_raw <- readr::read_csv(here::here("data", "1900s", "MENDENHALL_DATA (1).csv")) %>% 
  clean_names()
```

### 3---USGS Groundwater Wells 


```{r}
# Read in the full dataset for all well water measurements for the USA 

entire_usa_gw <- readr::read_tsv(here::here("data", "entire_usa_2021.txt")) 

```


# Tidy Data 

- Lowercase/clean names
- Remove extra blank columns 
- Remove values without well depth value/depth is zero & no static water level 
- Remove values without a lat/long 
- Rename columns to be consistent with other data frames 
- Add artesian conditions/columns 


### 1-- Early 1900s Jasechko Lab 

```{r}
# Join two dataframes to incorporate adjusted lat/long 

joined_jl <- jl_early1900s_raw %>% 
  full_join(latlong_adjust, by = "id")

# Tidy data frame 

early_1900s_tidy <- joined_jl %>% 
  clean_names() %>% 
  select(-x20:-x46, -citation_y, -state_y) %>% 
  filter(!is.na(max_depth_ft),
         !is.na(latitude)) %>% 
  filter(!(is.na(max_height_water_ft) & is.na(height_water_text)), 
         !height_water_text %in% c("+-", "+", "-")) %>% 
  rename(well_depth_ft = max_depth_ft, 
         water_level_ft = max_height_water_ft) %>% 
  mutate(artesian = ifelse(water_level_ft >= 0 | !is.na(height_water_text), 1, 0)) %>% 
  mutate(latitude_adj = ifelse(!is.na(new_lat), new_lat, latitude), 
         longitude_adj = ifelse(!is.na(new_lon), new_lon, longitude)) %>% 
  select(-latitude, -longitude, -new_lat:-diff_lon) %>% 
  rename(latitude = latitude_adj, 
         longitude = longitude_adj)

early_1900s_tidy$id<- sub("^", "w", early_1900s_tidy$id)


```

### 2-- Mendenhall et al., 1916

```{r}
mendenhall_tidy <- mendenhall_raw %>% 
  rename(id = site_id, 
         latitude = lat_dd, 
         longitude = long_dd, 
         water_level_ft = water_depth_ft) %>% 
  mutate(correct_water_levelft = water_depth_m*3.28084, 
           artesian = ifelse(correct_water_levelft == 0 | water_level_ft == 0, 1, 0))

mendenhall_tidy$id <- sub("^", "w", mendenhall_tidy$id)
```

### 3-- USGS Groundwater Wells 

(Guess that negative number indicates artesian condition--emailed USGS to be sure: correct, neg = artesian, pos = below land surface)

```{r}
# Entire USGS groundwater well data until fall 2019 (last updated) 

usa_early <- entire_usa_gw %>% 
  filter(!well_depth_va == "NA" | !hole_depth_va == "NA") %>% 
  filter(!well_depth_va == 0.00) %>% 
  filter(!lev_dt == "NA") %>% 
  filter(!level_year >1906) %>% 
  filter(!is.na(lev_va))

# Find median water level for wells that have more than one measurement per year 

early_usgs_med <- usa_early %>% 
  group_by(site_no) %>% 
  mutate(median = median(lev_va)) %>% 
  distinct(site_no, .keep_all = TRUE) %>% 
  rename(id = site_no, 
         latitude = dec_lat_va, 
         longitude = dec_long_va, 
         well_depth_ft = well_depth_va, 
         water_level_ft = lev_va) %>% 
  mutate(artesian = ifelse(water_level_ft <= 0, 1, 0))

early_usgs_med$id <- sub("^", "w", early_usgs_med$id)
  
```


# Prepare for Analysis 

1. Join data frames (id, lat, long, well depth, water level) 
2. Find deep wells (100 m/328 ft or greater)
3. Find deep wells that are also artesian 
4. Export to ArcGIS 

```{r}
# Join data frames 
# First tidy for join 

join_jl <- early_1900s_tidy %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)

join_mendenhall <- mendenhall_tidy %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)

join_usgs <- early_usgs_med %>% 
  select(id, latitude, longitude, well_depth_ft, water_level_ft, artesian)
```

```{r}
# Join data frames 

early1900s_all <- rbind(join_jl, join_mendenhall, join_usgs)
```

```{r}
# Tidy data 
# Remove data without well depth 
# Remove data without artesian indication 

early1900s_arc <- early1900s_all %>% 
  filter(!is.na(artesian)) 
  
```


Write as tsv for ArcGIS 

```{r}
# Write now-cleaned data as text file for ArcGIS

write_tsv(early1900s_arc, here::here("output_data", "early1900s_arc.txt"))
```

