---
title: "Flipped VHG"
author: "Annette Hilton"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Attach Packages

```{r}
# Attach packages 

library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(broom)
library(tibble)
library(dplyr)
library(tidyr)
library(ggpubr)
library(data.table)

# Disable scientific notation 

options(scipen=999)
```

# Methods & Data Sources

## Read in data 

Specify column type for "arc_artesian" to avoid NA values 

```{r}

spear_modern <- readr::read_csv(here::here("output_data", "spear_modernzones.txt"))

wells_total <- readr::read_tsv(here::here("output_data", "wells_total_n.txt")) %>% 
  distinct(id, .keep_all = TRUE)
```

## Data Analysis 

1. Join tables by township range 
2. Determine if VHG flipped 

```{r}
# join dataframes 
# tidy & rename columns 

joined_usa <- wells_total %>% 
  left_join(spear_modern, by = "id")

# mutate new columns for spearman correlation being "up" or "down" 
# mutate new column for flipped/change 

flipped <- joined_usa %>% 
  mutate(flipped = case_when(artesian == 1 & artesian_id == "artesian" ~"still_artesian",
                             artesian == 0 & artesian_id == "artesian" ~"pressurized",
                             artesian == 1 & estimate > 0 ~"still_pressurized",
                             artesian == 1 & estimate < 0 ~"depressurized",
                             artesian == 0 & estimate < 0 ~"still_depressurized", 
                             artesian == 0 & estimate > 0 ~"pressurized"))


```

Extra

```{r}
down_shift <- flipped_distinct %>% 
  filter(flipped == "down_shift")

still_up <- flipped_distinct %>% 
  filter(flipped == "still_up")

up_shift <- flipped_distinct %>% 
  filter(flipped == "upshift")

still_down <- flipped_distinct %>% 
  filter(flipped == "still_down")

originally_up <- flipped_distinct %>% 
  filter(vhg_1900s == "up" & !is.na(flipped))

originally_down <- flipped_distinct %>% 
  filter(vhg_1900s == "down")
```


 Write to csv file 
 
```{r}
write_csv(flipped, here::here("output_data", "flipped.txt"))
```

Arc Instructions: 

1. Import datafiles 
2. Join text files with PLSS file 
3. Create shapefiles individual (1900s and 2000s) 
4. Dissolve shapefile by township ra
5. Join with original file 

