---
title: "Modern Well Records (~2010-2020)"
author: "Annette Hilton"
date: "05/10/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Continental USA modern data & analysis 

## Attach Packages

```{r}
# Attach packages 

library(plyr)
library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(dplyr)
library(tidyr)
library(naniar)
library(lubridate)


# Disable scientific notation 

options(scipen=999)
```

### Read in Data

```{r}
# Read in the full dataset for all well water measurements for the USA 

entire_usa_gw <- readr::read_tsv(here::here("data", "entire_usa_2022.txt"))
```
test 

```{r}
f_status <- entire_usa_gw %>% 
  filter(lev_status_cd == "F") %>% 
  filter(level_year >= 2010) %>%
  distinct(site_no, .keep_all = T)

write_tsv(f_status, here::here("output_data", "f_status.txt"))

p_status <- entire_usa_gw %>% 
  filter(lev_status_cd == "P") %>%
  filter(level_year >= 2010) %>%
  distinct(site_no, .keep_all = T)

write_tsv(p_status, here::here("output_data", "p_status.txt"))
```

### Tidy Data 

- Remove wells without well depth 
- Remove wells with zero depth 
- Remove wells without a water level 
- Keep records only from 2010 and later 
- Calculate median water level for that time period 
- Indicate artesian conditions
 
 
 2010-2021 
 
```{r}
usgs_modern <- entire_usa_gw %>% 
  filter(!well_depth_va == "NA") %>% 
  filter(!well_depth_va == 0.00) %>% 
  filter(!lev_dt == "NA") %>%
  filter(!is.na(lev_va)) %>% 
  filter(level_year >= 2010)
  

usgs_modern_med <- usgs_modern %>% 
  group_by(site_no) %>% 
  mutate(median = median(lev_va)) %>% 
  distinct(site_no, .keep_all = TRUE) %>% 
  mutate(artesian = ifelse(median <= 0, 1, 0)) %>% 
  select(site_no, dec_lat_va, dec_long_va, well_depth_va, median, level_date, level_year, id.x, artesian)

usgs_modern_med$site_no <- sub("^", "w", usgs_modern_med$site_no)

write_tsv(usgs_modern_med, here::here("output_data", "usgs_modern.txt"))
```

quick test 05/18/22

```{r}
lev_status <- usgs_modern %>% 
  filter(!is.na(lev_status_cd))
```

### Add DEM file and Confined/Unconfined USGS Definitions file  

```{r}
# Modern data 
# Read in modern USGS data w/ DEM 

na_values <- c("-9999")

dem_usgs_modern <- readr::read_csv(here::here("data", "dem", "dem_usgs_modern.csv"), na = na_values)

```

```{r}
# Modern data 
# USGS confined/unconfined definitions and aquifer designations
# Data obtained by Dr. Scott Jasechko 4/20/2022 

usgs_confined_unconf <- readr::read_tsv(here::here("data", "confined_unconf_usgs_fromscott.txt")) %>% 
  clean_names() %>% 
  filter(!code == "NaN") %>% 
  mutate(site_no = sub("^", "w", site_no)) %>% 
  distinct(site_no, .keep_all = TRUE)

# Second pull of confined/unconfined data (04/25/2022) 

usgs_confined_unconf_2 <- readr::read_csv(here::here("data", "usgs_c_uc_wells_042522.txt")) %>% 
  clean_names() %>% 
  rename(national_aquifer = national_a, 
         local_aquifer = local_aqui, 
         dec_long_va = dec_long_v)

# Tidy above file for join (fix the site_no) 

usgs_clean <- usgs_confined_unconf_2 %>% 
  separate(site_no, c("name", "site_no")) %>% 
  mutate(site_no = sub("^", "w", site_no)) %>% 
  distinct(site_no, .keep_all = TRUE)

```

```{r}
# Join the two confined unconfined files from USGS 

joined_c_uc <- usgs_clean %>% 
  full_join(usgs_confined_unconf, by = c("site_no", "code", "dec_lat_va", "dec_long_va")) %>% 
  distinct(site_no, .keep_all = TRUE)

# Tidy file 

joined_tidy <- joined_c_uc %>% 
  select(site_no, national_aquifer.x, local_aquifer.x, code, dec_lat_va, dec_long_va, depth_m.x)

```


```{r}
# Join DEM file

modern_intermediate <- usgs_modern_med %>% 
  full_join(dem_usgs_modern, by = "site_no")

# Join confined/unconfined definitions file 

modern_final <- modern_intermediate %>% 
  left_join(joined_tidy, by = "site_no") %>% 
  select(-dec_lat_va.y, -dec_long_va.y)

write_tsv(modern_final, here::here("data", "modern_final.txt"))

```

# Individual Aquifer Systems 

### Mississippi Embayment 

```{r}
# Missouri, Illinois, Kentucky, Tennessee, Arkansas, Mississippi, Alabama, Louisiana

me_modern <- modern_final %>% 
  filter(id.x %in% c("al", "ar", "il", "ky", "la", "ms", "mo", "tn"))

write_tsv(me_modern, here::here("data", "me", "me_modern.txt"))
```

### North Atlantic Coastal Plain 

```{r}
# New York, New Jersey, Delaware, Maryland, Virginia, North Carolina 

nacp_modern <- modern_final %>% 
  filter(id.x %in% c("ny", "nj", "de", "md", "va", "nc")) 

write_tsv(nacp_modern, here::here("data", "nacp", "nacp_modern.txt"))
```

### Central Valley (California) 

USGS Data (above) 

```{r}
# USGS data 
cv_modern <- modern_final %>% 
  filter(id.x == "ca") %>% 
  clean_names()
```

Additional data from GAMA CA 

1. Upload data from Dr. Scott Jasechko (05/11/2022) of CA GAMA wells 
2. Tidy data, write reduced file for GIS DEM analysis 
3. DEM analysis 
4. Import back into R, join with USGS file 

```{r}
# Step 1. 
# GAMA CA Wells 

# Create path for multiple files 

gama_files = list.files(path = here::here("data", "cv", "additional_ca_data"), pattern = "*.txt", full.names = TRUE)

# Read in all files to one dataframe using list from above and `ldply()`

na_values <- c("-1")

allgama_files = ldply(gama_files, read_tsv, na = na_values, show_col_types = FALSE)

```

```{r}
# Step 2. 
# Initial tidy

ca_gama_tidy <- allgama_files %>% 
  clean_names() %>% 
  filter(!depth <= 0) %>% 
  filter(!water_level < -5000) %>% 
  mutate(date = lubridate::parse_date_time(datetime, orders = c("ymd", "ymd HMS"))) %>% 
  mutate(year = lubridate::year(date)) %>% 
  filter(year >= 2010)
  
# Calculate median, artesian 

ca_gama_tidy2 <- ca_gama_tidy %>% 
  group_by(well_id) %>% 
  mutate(median = median(water_level)) %>% 
  distinct(well_id, .keep_all = TRUE) %>% 
  mutate(artesian = ifelse(median <= 0, 1, 0))

ca_gama_tidy2$well_id<- sub("^", "w", ca_gama_tidy2$well_id)

# write to tsv for ArcGIS DEM processing 

write_tsv(ca_gama_tidy2, here::here("output_data", "cv", "ca_gama_demtoarc.txt"))

```

Step 3 (in ArcGIS: on external hard drive "look busy", file called `cv_dem_gama`)

```{r}
# Step 4. 
# Read in processed DEM data

cv_dem <- readr::read_csv(here::here("data", "cv", "dem_fromarc0513.txt")) %>% 
  clean_names()

# Make consistent with USGS data to join them 
ca_dem_tidy <- cv_dem %>% 
  mutate(id_x = "ca", 
         national_aquifer_x = NA, 
         local_aquifer_x = NA, 
         code = NA,
         depth_m_x = NA) %>% 
  rename(site_no = well_id, 
         dec_lat_va_x = lat, 
         dec_long_va_x = lon, 
         well_depth_va = depth, 
         level_date = date,
         level_year = year, 
         rastervalu = cali) %>% 
  select(site_no, dec_lat_va_x, dec_long_va_x, well_depth_va, median, level_date, level_year, 
         id_x, artesian, rastervalu, national_aquifer_x, local_aquifer_x, code, depth_m_x)
```


```{r}
# Join USGS and drilling well data 

cv_modern_final <- rbind(ca_dem_tidy, cv_modern)

write_tsv(cv_modern_final, here::here("data", "cv", "cv_modern.txt"))
```
 
### Floridan Aquifer System 

```{r}
# Florida, Georgia, South Carolina, Alabama

fl_modern <- modern_final %>% 
   filter(id.x %in% c("al", "fl", "ga", "sc")) 

write_tsv(fl_modern, here::here("data", "fl", "fl_modern.txt"))
```

### Texas Gulf Coast Aquifer System (Texas)

```{r}
# Texas

tx_modern <- modern_final %>% 
   filter(id.x == "tx") 

write_tsv(tx_modern, here::here("data", "tx", "tx_modern.txt"))
```

### Columbia Plateau (Washington State)

```{r}
# Washington State 

columbia_modern <- modern_final %>% 
   filter(id.x == "wa") 

write_tsv(columbia_modern, here::here("data", "columbia", "columbia_modern.txt"))
```

### Roswell Artesian Basin (New Mexico) 

```{r}
# New Mexico 

roswell_modern <- modern_final %>% 
   filter(id.x == "nm") 

write_tsv(roswell_modern, here::here("data", "roswell", "roswell_modern.txt"))
```

### Dakota Aquifer (South Dakota) 

USGS Data (above) 

```{r}
sd_modern <- modern_final %>% 
  filter(id.x == "sd") %>% 
  clean_names()
```

Additional data from SD Observation Wells 

1. Import data of Observation wells from South Dakota Department of Agriculture and Natural Resources (https://apps.sd.gov/nr69obswell/default.aspx#viewHelp). All records up to 11/18/2021
2. Tidy data, write for ArcGIS Pro DEM analysis 
3. DEM analysis 
3. Import back into R, join with USGS file  


```{r, results=FALSE, message=FALSE}
# Step 1. 
# Well Readings--water level, dates 

sd_wells_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "well_readings"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))

# Metadata--well depth 

sd_depth_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "meta_data"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))
```


```{r}
# Step 2. 
# Tidy data, make consistent with USGS format 
# Correct water level (SD Obs wells indicate artesian values w/ positive number, below land surface with negative)

# Join data together 
sd_join <- sd_wells_raw %>% 
  full_join(sd_depth_raw, by = "Observation Well")

# Tidy data 

sd_full_tidy <- sd_join %>% 
  clean_names() %>% 
  select(-is_manual, -x8, -x14) %>% 
  mutate(date = lubridate::mdy(reading_date), 
         year = year(date)) %>% 
  filter(year >= 2010) %>% 
  filter(!is.na(reading)) %>% 
  filter(!is.na(total_depth)) 

sd_full_tidy$reading <- as.numeric(sd_full_tidy$reading)

sd_full_med <- sd_full_tidy %>% 
  mutate(corrected_water_lev = (reading*-1)) %>% 
  group_by(observation_well) %>% 
  mutate(median = median(corrected_water_lev)) %>% 
  distinct(observation_well, .keep_all = TRUE) %>% 
  mutate(artesian = ifelse(median <= 0, 1, 0))

# Write for ArcGIS DEM analysis 
write_tsv(sd_full_med, here::here("output_data", "sd", "sd_obs_wells_dem.txt"))

```

Step 3 (in ArcGIS: on external hard drive "look busy", file called `south_dakota`)

```{r}
# Step 4. 
# Read in processed DEM data

sd_dem <- readr::read_csv(here::here("data", "south_dakota", "sd_dem_051322.txt")) %>% 
  clean_names()

# Make consistent with other data frame 

sd_obs_join <- sd_dem %>% 
  mutate(id_x = "sd", 
         national_aquifer_x = NA, 
         local_aquifer_x = NA, 
         code = NA,
         depth_m_x = NA) %>%
rename(site_no = observation_well, 
         dec_lat_va_x = latitude, 
         dec_long_va_x = longitude, 
         well_depth_va = total_depth, 
         level_date = date,
         level_year = year, 
         rastervalu = ned10m_bilinear_cusa_albers102003 ) %>% 
  select(site_no, dec_lat_va_x, dec_long_va_x, well_depth_va, median, level_date, level_year, 
         id_x, artesian, rastervalu, national_aquifer_x, local_aquifer_x, code, depth_m_x)
```

```{r}
# Join USGS and drilling well data 

sd_modern_final <- rbind(sd_obs_join, sd_modern)

write_tsv(sd_modern_final, here::here("data", "south_dakota", "sd_modern.txt"))
```
 
 
# Depth to Confined Investigations (05/18/22) 

1. Spatial join modern data (`modern_final.txt` from above) with aquifer file and depth data: processsed in ArcPro (file name: ) 
2. Load into R for analysis 

```{r}
# Step 2. 
modern_depths <- readr::read_csv(here::here("data", "depths", "modern_depth_joined.txt"))
```

```{r}
# How many wells do we have for each aquifer system that is greater than the confined depth cut off? 

modern_analysis_depth <- modern_depths %>% 
  mutate(depth_m = well_depth_va * 0.3048) %>% 
  mutate(confined_depth = ifelse(depth_m >= DepthConfi, "confined depth", NA))

modern_depth_final <- modern_analysis_depth %>% 
  filter(!is.na(confined_depth)) %>% 
  group_by(Aquifer_1) %>% 
  filter(n() >= 4)

write_tsv(modern_depth_final, here::here("data", "depths", "modern_depth_final.txt"))
```

How many aquifers? 

```{r}
aquifer_modern <- modern_depth_final %>% 
  distinct(Aquifer_1, .keep_all = T)
```

Join two 

```{r}
both_hist_modern <- aquifer_modern %>% 
  inner_join(aquifer_historical, by = "Aquifer_1")

write_tsv(both_hist_modern, here::here("data", "depths", "both_aquiferstimeperiod.txt"))
```

