---
title: "Results South Dakota--Change over time between early 1900s wells and modern conditions (2010-2020)"
author: "Annette Hilton"
date: "11/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Attach Packages

```{r}
# Attach packages 

library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(broom)
library(tibble)
library(dplyr)
library(tidyr)
library(ggpubr)

# Disable scientific notation 

options(scipen=999)
```

# South Dakota 

## Change over time between early 1900s data and modern (2010-2020) data 

Methods: 

1. Compilation of early well data (see "early_1900s_cleaning.Rmd"--deep wells) and modern data for South Dakota (drilling wells, and observation monitoring wells) (see "sd_modern_data_cleaning.Rmd")

Note: Observation wells (negative = below land surface, positive = artesian) 
Drilling wells = all positive (below land surface) 

2. Exported into ArcGIS

3. Buffer of 10 km around deep early 1900s wells, intersect with modern wells 

4. Import into R for data wrangling 

5. Data cleaning & tidying, prepare for statistical tests 

6. Spearman rank correlations 


## Import Data 

### Methods step 4. 

```{r}
# Read in dataset of intersect between early 1900s wells and modern wells within 10 km distance 

intersect_sd <- readr::read_csv(here::here("data", "south_dakota", "sd_intersect_alldates.txt"))

```

## Data cleaning, tidy, prepare for stats 

### Methods step 5. 

```{r}

intersect_tidy <- intersect_sd %>% 
  clean_names() %>% 
  select(id, latitude, longitude, well_depth, water_leve, artesian, id_1, depth, median_lev, artesian_1, lat, long, year)
  
```

Add criteria: 

1. Each original 1900s well must have at least 3 modern wells associated 

2. Make sure each original well has at least one well that is more shallow and one well that is deeper than it 

3. Remove artesian wells if the artesian conditions are more than 50% of all wells in the area zone 
 

```{r}
# 1. Criteria of 3 wells per early 1900s well 
# 2. 1900s must have a modern well deeper and shallower than it 

 wells_criteria <- intersect_tidy %>% 
  group_by(id) %>% 
  filter(n() >= 3) %>% 
  mutate(min_depth = min(depth), 
         max_depth = max(depth) + 20) %>% 
  filter(well_depth < max_depth & well_depth > min_depth) %>% 
  ungroup()

wells_criteria_dates <- wells_criteria %>% 
  filter(year >= 2010)

```

## Graph Well Depths Over Time 

Graph individual zones 

```{r}


# Use group_split() to indicate separation by id

plot_vhgzones<- wells_criteria_depth %>%
  group_split(id)

# Make a function to call each ID and spearman estimate as the title of the graph

plots <- function(data){

  plot_name_well <- data$id[1]

ggplot(data, aes(x = year, y = depth)) +
  geom_point() +
  labs(x = "Depth (ft)",
       y = "Water Level (ft)",
       title = plot_name_well, 
       subtitle = plot_name) +
  scale_y_reverse()
}


all_plots <- purrr::map(plot_vhgzones, plots)

all_plots[[1]]

# Save as PDF

pdf(here::here("output_data", paste0("vhg_spear_sd", Sys.Date(), ".pdf")), height = 11, width = 8.5)
ggarrange(plotlist = all_plots, ncol = 1, nrow = 3, align = "v")
dev.off()
```


```{r}
# 3. Artesian wells 

# Group by ID 
# Add counts for total number of wells in township
# Add counts for total number of artesian wells in each township 
# Add percentage of artesian wells per township 

wells_total_n <- wells_criteria %>% 
  group_by(id) %>% 
  add_count(id) %>% 
  add_tally(artesian_1 == 1, name = "art_n") %>% 
  mutate(percent = (art_n/n)) %>% 
  mutate(artesian_id = ifelse(percent > 0.5, "artesian", NA))

write_tsv(wells_total_n, here::here("output_data", "wells_total_n.txt"))

# Add criteria of 50% or greater artesian are removed 

no_artesian <- wells_total_n %>% 
  group_by(id) %>% 
  filter(!percent > 0.5)

# Keep all artesian well townships 

all_artesian <- wells_total_n %>% 
  group_by(id) %>% 
  filter(percent > 0.5)

# # Write artesian file for Arc 
# 
# artesian_arc <- all_artesian %>% 
#   mutate(arc_artesian = "artesian") %>% 
#   distinct(id, .keep_all = TRUE)
# 
# write_csv(artesian_arc, here::here("output_data", "artesian_arc.txt"))

```

## Statistical tests 

### Methods step 6. 

Spearman rank correlation method calculates the correlation between the ranks of x and the ranks of y variables. 

Variable: Depth of well (well_depth)
Variable: Water level (water_leve)


IMPORTANT NOTE: Something is going on with cor.test(), does not work. cor() does work. Error: "not enough finite values" Have thoroughly investigated and do not understand. 

```{r}
# Use lapply() and split() to split the dataframe by "id" (original early 1900s well) and then iterate the cor.test function for each group (one original 1900s well and modern wells associated)

# spearman <- lapply(split(no_artesian, f = no_artesian$id), function(no_artesian) cor.test(no_artesian$med_level, no_artesian$well_dep_1, method = "spearman", exact = FALSE))

spearman <- lapply(split(no_artesian, f = no_artesian$id), function(no_artesian) cor(no_artesian$median_lev, no_artesian$depth, method = "spearman"))

spearman_2010 <- lapply(split(wells_criteria_dates, f = wells_criteria_dates$id), function(wells_criteria_dates) cor(wells_criteria_dates$median_lev, wells_criteria_dates$depth, method = "spearman"))

# Bind results together in a dataframe and indicate the id (rownames_to_column())

results <- as.data.frame(do.call(rbind, spearman)) %>% 
  rownames_to_column(var = "id") %>% 
  rename(estimate = V1)

results_2010 <- as.data.frame(do.call(rbind, spearman_2010)) %>% 
  rownames_to_column(var = "id") %>% 
  rename(estimate = V1)

# 
# results <- as.data.frame(do.call(rbind, spearman)) %>% 
#   rownames_to_column(var = "id") %>% 
#   select(id, p.value, estimate)

# Remove character string in estimate column (rho)
# Make sure p-value is numeric 

# results$estimate <- gsub("[a-z = ()]", "", results$estimate)
# results$p.value <- as.numeric(results$p.value)

# Write as csv 

write_csv(results, here::here("output_data", "spear_modernzones.txt"))

```

## Graph 

Graph individual zones 

```{r}

# Join results of spearman ranks with full dataset 

full_data <- no_artesian %>% 
  inner_join(results, by = "id")

write_tsv(full_data, here::here("output_data", "sd_spear_full.txt"))

full_data_2010 <- wells_criteria_dates %>% 
  inner_join(results_2010, by = "id")

write_tsv(full_data, here::here("output_data", "sd_spear_full_2010.txt"))

# Use group_split() to indicate separation by id

plot_vhgzones<- full_data_2010 %>%
  group_split(id)

# Make a function to call each ID and spearman estimate as the title of the graph

plots <- function(data){

  plot_name <- data$estimate[1]
  plot_name_well <- data$id[1]

ggplot(data, aes(x = depth, y = median_lev)) +
  geom_point() +
  labs(x = "Depth (ft)",
       y = "Water Level (ft)",
       title = plot_name_well, 
       subtitle = plot_name) +
  scale_y_reverse()
}


all_plots <- purrr::map(plot_vhgzones, plots)

all_plots[[1]]

# Save as PDF

pdf(here::here("output_data", paste0("vhg_spear_sd", Sys.Date(), ".pdf")), height = 11, width = 8.5)
ggarrange(plotlist = all_plots, ncol = 1, nrow = 3, align = "v")
dev.off()
```


----------------------------------------

USGS DATA 


## Import Data 

### Methods step 4. 

```{r}
# Read in dataset of intersect between early 1900s wells and modern wells within 10 km distance 

intersect_sd_usgs <- readr::read_csv(here::here("data", "south_dakota", "sd_allwells_intersect.txt"))


```

## Data cleaning, tidy, prepare for stats 

### Methods step 5. 

```{r}

intersect_usgs <- intersect_sd_usgs %>% 
  clean_names() %>% 
  select(id, latitude, longitude, well_depth, water_leve, artesian, site_no, level_year, well_dep_1, lev_va, dec_lat_va, dec_long_v)
  

```

Add criteria: 

1. Each original 1900s well must have at least 3 modern wells associated 

2. Make sure each original well has at least one well that is more shallow and one well that is deeper than it 

3. Remove artesian wells if the artesian conditions are more than 50% of all wells in the area zone 
 
```{r}
# USGS 

 wells_criteria_usgs <- intersect_usgs %>% 
   filter(level_year <= 1960 & level_year >= 1950) %>% 
   group_by(site_no) %>% 
   mutate(med_level = median(lev_va)) %>% 
   distinct(site_no, .keep_all = TRUE) %>% 
   ungroup() %>% 
   group_by(id) %>% 
   filter(n() >= 3) %>% 
  mutate(min_depth = min(well_dep_1), 
         max_depth = max(well_dep_1)) %>% 
  filter(well_depth < max_depth & well_depth > min_depth) %>% 
  ungroup()

```

USGS 

```{r}
# 3. Artesian wells 

# Group by ID 
# Add counts for total number of wells in township
# Add counts for total number of artesian wells in each township 
# Add percentage of artesian wells per township 

wells_total_n_usgs <- wells_criteria_usgs %>% 
  mutate(artesian_usgs = ifelse(med_level <= 0, 1, 0)) %>% 
  group_by(id) %>% 
  add_count(id) %>% 
  add_tally(artesian_usgs == 1, name = "art_n") %>% 
  mutate(percent = (art_n/n)) %>% 
  mutate(artesian_id = ifelse(percent > 0.5, "artesian", NA))

write_tsv(wells_total_n, here::here("output_data", "wells_total_n.txt"))

# Add criteria of 50% or greater artesian are removed 

no_artesian <- wells_total_n_usgs %>% 
  group_by(id) %>% 
  filter(!percent > 0.5)

no_artesian$med_level <- as.numeric(no_artesian$med_level)
no_artesian$well_dep_1 <- as.numeric(no_artesian$well_dep_1)

# Keep all artesian well townships 

all_artesian <- wells_total_n %>% 
  group_by(id) %>% 
  filter(percent > 0.5)

# # Write artesian file for Arc 
# 
# artesian_arc <- all_artesian %>% 
#   mutate(arc_artesian = "artesian") %>% 
#   distinct(id, .keep_all = TRUE)
# 
# write_csv(artesian_arc, here::here("output_data", "artesian_arc.txt"))

```

## Statistical tests 

### Methods step 6. 

Spearman rank correlation method calculates the correlation between the ranks of x and the ranks of y variables. 

Variable: Depth of well (well_depth)
Variable: Water level (water_leve)


IMPORTANT NOTE: Something is going on with cor.test(), does not work. cor() does work. Error: "not enough finite values" Have thoroughly investigated and do not understand. 

USGS 

```{r}
# Use lapply() and split() to split the dataframe by "id" (original early 1900s well) and then iterate the cor.test function for each group (one original 1900s well and modern wells associated)

spearman_usgs <- lapply(split(no_artesian, f = no_artesian$id), function(no_artesian) cor(no_artesian$well_dep_1, no_artesian$med_level, method = "spearman"))

# Bind results together in a dataframe and indicate the id (rownames_to_column())

results <- as.data.frame(do.call(rbind, spearman_usgs)) %>% 
  rownames_to_column(var = "id") %>% 
  rename(estimate = V1)
# 
# results <- as.data.frame(do.call(rbind, spearman)) %>% 
#   rownames_to_column(var = "id") %>% 
#   select(id, p.value, estimate)

# Remove character string in estimate column (rho)
# Make sure p-value is numeric 

# results$estimate <- gsub("[a-z = ()]", "", results$estimate)
# results$p.value <- as.numeric(results$p.value)

# Write as csv 

write_csv(results, here::here("output_data", "spear_southdakota.txt"))

```


## Graph 

Graph individual zones 

```{r}

# Join results of spearman ranks with full dataset 

full_data <- no_artesian %>% 
  inner_join(results, by = "id")

write_tsv(full_data, here::here("output_data", "sd_spear_full.txt"))

# Use group_split() to indicate separation by id

plot_vhgzones<- full_data %>%
  group_split(id)

# Make a function to call each ID and spearman estimate as the title of the graph

plots <- function(data){

  plot_name <- data$estimate[1]
  plot_name_well <- data$id[1]

ggplot(data, aes(x = well_dep_1, y = med_level)) +
  geom_point() +
  labs(x = "Depth (ft)",
       y = "Water Level (ft)",
       title = plot_name_well, 
       subtitle = plot_name) +
  scale_y_reverse()
}


all_plots <- purrr::map(plot_vhgzones, plots)

all_plots[[1]]

# Save as PDF

pdf(here::here("output_data", paste0("vhg_spear_sd", Sys.Date(), ".pdf")), height = 11, width = 8.5)
ggarrange(plotlist = all_plots, ncol = 1, nrow = 3, align = "v")
dev.off()
```


