---
title: "South Dakota: Modern Well Records (~2010-2020)"
author: "Annette Hilton"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# South Dakota Modern Analysis 

## Attach Packages

```{r}
# Attach packages 

library(parzer)
library(Rcpp)
library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(broom)
library(tibble)
library(dplyr)
library(tidyr)
library(ggpubr)
library(data.table)
library(lubridate)


# Disable scientific notation 

options(scipen=999)
```

# Methods 

1. Read in modern data of South Dakota 
  a. Drilling wells from South Dakota Department of Agriculture and Natural Resources (https://apps.sd.gov/nr68welllogs/). From 1/01/1800-11/22/2021 
  b. Observation wells from South Dakota Department of Agriculture and Natural Resources (https://apps.sd.gov/nr69obswell/default.aspx#viewHelp). All records up to 11/18/2021
2. Tidy data 
3. Prepare for Arc 

# Read in Data

```{r}
sd_raw_drill <-
    list.files(path = here::here("data", "south_dakota", "sd_drill_wells"),
               pattern = "*.csv", 
               full.names = T) %>% 
    map_df(~read_csv(., col_types = cols(.default = "c"))) 

```

Drilling well depths over time 

```{r}

sd_drill_depths <- sd_raw_drill %>% 
  clean_names() %>% 
  select(-guid, -comp_rpt_pdf) %>% 
  mutate(date = lubridate::mdy(well_comp_date), 
         year = lubridate::year(date)) %>% 
  filter(!is.na(depth)) %>% 
  filter(!depth %in% c("9999", "9999.0", "999", "999.0", "0", "0.00", "0.0"))

sd_drill_depths$depth <- as.numeric(sd_drill_depths$depth)

# Make a plot of well depth over time 

well_depth_sd <- ggplot(sd_drill_depths, 
                        aes(x = year, 
                            y = depth)) +
  geom_line()

well_depth_sd

```

Histogram of depths 

```{r}
hist(sd_drill_depths$depth)
```


```{r}
# Tidy data 

sd_tidy_drill <- sd_raw_drill %>% 
  clean_names() %>% 
  select(-guid, -comp_rpt_pdf) %>% 
  mutate(date = lubridate::mdy(well_comp_date), 
         year = lubridate::year(date)) %>% 
  filter(!is.na(static_value), 
         !is.na(depth)) %>% 
  mutate(id = seq(1:2643), 
         artesian = ifelse(static_value <= 0, 1, 0)) # Apparently no artesian values? 

sd_tidy_drill$id <- sub("^", "w", sd_tidy_drill$id) 

# Make consistent with other data frame 

sd_drill_join <- sd_tidy_drill %>% 
  rename(original_date = well_comp_date, 
         median_level = static_value) %>% 
  mutate(elevation = NA) %>% 
  select(id, lat, long, aquifer, elevation, original_date, date, year, depth, median_level, artesian) 

sd_drill_join$median_level <- as.numeric(sd_drill_join$median_level)

```

## Read in Data 

Observation Wells From: https://apps.sd.gov/nr69obswell/default.aspx#viewHelp
Pos = artesian 
Neg = below land surface 

```{r}
# Well Readings--water level, dates 

sd_wells_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "well_readings"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))

# Metadata--well depth 

sd_depth_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "meta_data"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))
```

## Tidy Data 


```{r}
# Full join on observation wells 

sd_join <- sd_wells_raw %>% 
  full_join(sd_depth_raw, by = "Observation Well")

# Tidy data 

sd_full_tidy <- sd_join %>% 
  clean_names() %>% 
  select(-is_manual, -x8, -x14) %>% 
  mutate(date = lubridate::mdy(reading_date), 
         year = year(date)) %>% 
  filter(!is.na(reading)) %>% 
  filter(!is.na(total_depth))

sd_full_tidy$reading <- as.numeric(sd_full_tidy$reading)

sd_full_med <- sd_full_tidy %>% 
  group_by(observation_well) %>% 
  mutate(median_level = median(reading)) %>% 
  distinct(observation_well, .keep_all = TRUE) %>% 
  mutate(artesian = ifelse(median_level >= 0, 1, 0))


# Make consistent with other data frame 

sd_obs_join <- sd_full_med %>% 
  rename(original_date = reading_date, 
         water_level = reading, 
         lat = latitude, 
         long = longitude, 
         id = observation_well, 
         depth = total_depth) %>% 
  select(id, lat, long, aquifer, elevation, original_date, date, year, depth, median_level, artesian)

```

Join both dataframes 

```{r}
sd_full <- rbind(sd_obs_join, sd_drill_join)
```

Write for ArcGIS 

```{r}
write_tsv(sd_full, here::here("output_data", "sd_full_modern_dates.txt"))
```


