---
title: "Columbia Plateau--Well Classification"
author: "Annette Hilton"
date: "02/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the Columbia Plateau Aquifer System (Washington State) as compared to present day (2010-2021).

The purpose of this Rmd is to: 

1) Prep data for input to ArcGIS (for hydrogeological analysis & DEM): data prepped in `sd_modern_data_cleaning.Rmd` 
2) After ArcGIS processing: 

1) Determine well-aquifer classification 
2) Compare artesian conditions 
  

## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)

# Disable scientific notation 

options(scipen=999)
```

## ArcGIS Data Analysis 

The following steps were performed in ArcGIS: 

Hydrogeological data: 
Arcfile = "columbia" 

1. Hydrogeologic layers (raster files) 
2. Modern wells from Washington State (USGS) (processed in Rmd `sd_modern_data_cleaning.Rmd`) and early 1900s data (early1900s_dem_wgs84_final.txt)
3. Use tool "Extract multi values to points" to determine well location intersection with geological raster layers
4. Export resulting well data as a text files

## Data Analysis (Post-Arc)

### Hydrogeological data 

```{r}
# Read in modern data file from Arc 

cp_mod_fromarc <- readr::read_csv(here::here("data", "columbia", "mod_columbia_classify.txt"))

# Read in early 1900s data file from Arc 

cp_early1900s_fromarc <- read_csv(here::here("data", "columbia", "early1900s_columbia_classify.txt"))

```

### Tidy Data 

*Note:* You must join metadata files/original files for modern and early 1900s well information with the Arc processed files.
Arc somehow messes up numbers in its processing, so we cannot trust the well depth and well level information after it has been processed in Arc. 
Join by ID/site_no. 


Modern Data Joins 

Read in metadata files for modern USGS data (to get DEM) 

```{r}
# Modern data 
# Read in modern USGS data w/ DEM 
dem_usgs_modern <- readr::read_csv(here::here("data", "dem", "dem_usgs_modern.csv")) 

# Read in modern USGS data full file (from Rmd `modern_data_cleaning.Rmd`)
usgs_modern <- readr::read_tsv(here::here("output_data", "usgs_modern.txt"))

```

Join DEM files with full data files 
Join with processed hydrogeological data from Arc 

```{r}
# Join modern 

modern_join <- usgs_modern %>% 
  full_join(dem_usgs_modern, by = "site_no")

# Join with hydrogeological processed data from Arc 

columbia_join <- modern_join %>% 
  full_join(cp_mod_fromarc, by = "site_no") %>% 
  clean_names()

# Tidy modern data 
# Filter for just Washington state 

cp_mod_tidy <- columbia_join %>%  
  select(-objectid:-artesian_y) %>% 
  filter(id_x == "wa")

```

Early 1900s Data Joins 

Read in meta data files for early 1900s data 

```{r}
# Read in meta file for early 1900s 

early1900s_meta <- readr::read_tsv(here::here("output_data", "early1900s_arc.txt"))

```

Join DEM files with full data files 
Join with processed hydrogeological data from Arc 

```{r}
# 1900s data 

cp_early1900s_tidy <- cp_early1900s_fromarc %>% 
  clean_names() %>% 
  select(-objectid, -oid)

# Join with meta file 

cp_early1900s_join <- cp_early1900s_tidy %>% 
  full_join(early1900s_meta, by = "id") %>% 
  clean_names() %>% 
  select(-latitude_x:-artesian_x)

```


### Columbia Plateau Aquifer Information 

Hydrologic Model Layer Surfaces:

- Overburden (surficial) (obtop) 
- Saddle Mountains Basalt (smtop) 
- Wanapum Basalt (wntop) 
- Grande Ronde Basalt (grtop) 
- Older Bedrock (pmtop) 


### Aquifer analysis 

- Find well bottom 

```{r}
# Convert `rastervalu` column from meters to feet (conversion: 1 meter = 3.28084 feet)
# Add a column that is elevation minus well depth
# Add a column that is elevation minus well water level 

# Modern data 

cp_analysis_mod <- cp_mod_tidy %>% 
  mutate(topo_ft = rastervalu * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_va_x) %>% 
   mutate(keep = 
           ifelse(is.na(obtop_f) & is.na(smtop_f) & is.na(wntop_f) & is.na(grtop_f) & is.na(pmtop_f), "NA", "keep")) %>% 
  filter(!keep == "NA")

 
# 1900 data 

cp_analysis_1900 <- cp_early1900s_join %>% 
  mutate(topo_ft = ned10m_bilinear_cusa_albers102003 * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_ft_y) %>% 
  mutate(keep = 
           ifelse(is.na(obtop_f) & is.na(smtop_f) & is.na(wntop_f) & is.na(grtop_f) & is.na(pmtop_f), "NA", "keep")) %>% 
  filter(!keep == "NA")

```

#### Classify each well to an aquifer unit 

Theory: 

well bottom < top of Overburden (obtop) & well bottom > top of Saddle Mtns Basalt (smtop) = Overburden (ob) 
well bottom < top of Saddle Mtns Basalt (smtop) & well bottom > top of Wanapum Basalt (wntop) = Saddle Mountains Basalt (smb) 
well bottom < top of Wanapum Basalt (wntop) & well bottom > top of Grande Ronde Basalt (grtop) = Wanapum Basalt (wnb) 
well bottom < top of Grande Ronde Basalt (grtop) & well bottom > top of Older Bedrock (pmtop) = Grande Ronde Basalt (grb) 
well bottom < top of Older Bedrock = Older Bedrock (pmb) 


Top Only Analysis 

```{r}
# Classify wells into their aquifer units 

# Modern data 

cp_class_mod <- cp_analysis_mod %>% 
  rowwise() %>%
   mutate(aquifer = 
        ifelse(well_bottom <= obtop_f & well_bottom >= max(smtop_f, wntop_f, grtop_f, pmtop_f,
                                                          na.rm = TRUE), "ob", 
        ifelse(well_bottom <= smtop_f & well_bottom >= max(wntop_f, grtop_f, pmtop_f,
                                                          na.rm = TRUE), "smb", 
        ifelse(well_bottom <= wntop_f & well_bottom >= max(grtop_f, pmtop_f,
                                                          na.rm = TRUE), "wnb",
        ifelse(well_bottom <= grtop_f & well_bottom >= max(pmtop_f, na.rm = TRUE), "grtop", 
        ifelse(well_bottom <= pmtop_f, "pmb", NA))))))

# 1900 data 

cp_class_1900 <- cp_analysis_1900 %>% 
  rowwise() %>%
   mutate(aquifer = 
        ifelse(well_bottom <= obtop_f & well_bottom >= max(smtop_f, wntop_f, grtop_f, pmtop_f,
                                                          na.rm = TRUE), "ob", 
        ifelse(well_bottom <= smtop_f & well_bottom >= max(wntop_f, grtop_f, pmtop_f,
                                                          na.rm = TRUE), "smb", 
        ifelse(well_bottom <= wntop_f & well_bottom >= max(grtop_f, pmtop_f,
                                                          na.rm = TRUE), "wnb",
        ifelse(well_bottom <= grtop_f & well_bottom >= max(pmtop_f, na.rm = TRUE), "grtop", 
        ifelse(well_bottom <= pmtop_f, "pmb", NA))))))

```

Summary of wells per aquifer unit 

```{r}
# Summary of wells per aquifer unit 

# Modern data 
summarymod_cp <- cp_class_mod %>% 
  group_by(aquifer) %>% 
  count(aquifer) 


# 1900s data 

summary1900_cp <- cp_class_1900 %>% 
  group_by(aquifer) %>% 
  count(aquifer)

```

---------------------------------------------------------------
Confined criteria coded out for now
---------------------------------------------------------------

<!-- #### Confined Well Criteria  -->

<!-- Confined aquifer unit:  -->

<!-- - Criteria: Must have at least 10 m (33 ft) of confining material above -->
<!-- - Code: If the aquifer is confined, calculate the sum of the confining unit thickness above (layer4 & layer5). If the sum of the confining unit thickness is >= 33 ft (10m), then "keep" the well.  -->


<!-- First, determine thickness of aquifer units  -->

<!-- ```{r} -->
<!-- # Thickness of corcoran clay -->

<!-- # Modern data  -->

<!-- cv_thickness_mod <- cv_class_mod %>%  -->
<!--   mutate( -->
<!--     corcoran_thick = layer4_top - layer6_top) -->

<!-- # 1900 data  -->

<!-- cv_thickness_1900 <- cv_class_1900 %>%  -->
<!--   mutate( -->
<!--     corcoran_thick = layer4_top - layer6_top) -->

<!-- ``` -->

<!-- Then, determine cumulative thickness & which wells meet criteria  -->

<!-- ```{r} -->
<!-- # Modern data  -->

<!-- cv_cum_thickness_mod <- cv_thickness_mod %>%  -->
<!--   rowwise() %>%  -->
<!--   dplyr::mutate(keep =  -->
<!--            ifelse(corcoran_thick >= 33, "keep", NA)) -->

<!-- # Modern data  -->

<!-- cv_cum_thickness_1900 <- cv_thickness_1900 %>%  -->
<!--   rowwise() %>%  -->
<!--   dplyr::mutate(keep =  -->
<!--            ifelse(corcoran_thick >= 33, "keep", NA)) -->
<!-- ``` -->


#### Comparison of artesian conditions 

1900 to 2010-2020 

Modern n = 497
1900s n = 434

```{r}
# MODERN 

# Modern all 
modern_cp_all <- cp_class_mod 

# write file for arc 
write_tsv(modern_cp_all, here::here("output_data", "columbia", "modern_cp_all.txt"))

# Modern theoretically confined 
modern_cp_confined <- cp_class_mod %>% 
  filter(aquifer %in% c("grtop", "wnb", "pmb")) 

# write file for arc 
write_tsv(modern_cp_confined, here::here("output_data", "columbia", "modern_cp_confined.txt"))

# Modern theoretically unconfined 
modern_cp_unconf <- cp_class_mod %>% 
  filter(aquifer %in% c("ob", "smb")) 

# write file for arc 
write_tsv(modern_cp_unconf, here::here("output_data", "columbia", "modern_cp_unconf.txt"))


# 1900s 

# 1900s all 
historical_cp_all <- cp_class_1900 

# write file for arc 
write_tsv(historical_cp_all, here::here("output_data", "columbia", "historical_cp_all.txt"))

# 1900s confined 
historical_cp_confined <- cp_class_1900 %>% 
  filter(aquifer %in% c("grtop", "wnb", "pmb")) 

# write file for arc 
write_tsv(historical_cp_confined, here::here("output_data", "columbia", "historical_cp_confined.txt"))

# 1900s unconfined 
historical_cp_unconf <- cp_class_1900 %>% 
  filter(aquifer %in% c("ob", "smb")) 

# write file for arc 
write_tsv(historical_cp_unconf, here::here("output_data", "columbia", "historical_cp_unconf.txt"))
```

