---
title: "Change in Head by Aquifer System"
author: "Annette Hilton"
date: '2023-06-15'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Attach Packages

```{r}
# Attach packages 

library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(knitr)


# Disable scientific notation 

options(scipen=999)
```

# Read in data 

- By each regional aquifer system that you can analyze based on criteria (n=25 samples)
- Data that has already been sorted in aquifer unit, contains DEM information 
- Regional aquifer systems include: 
  - Mississippi Embayment
  - North Atlantic Coastal Plain 
  - Central Valley 
  - Dakota Aquifer System
  - Floridan 
  - Houston-Gulf Coast 
  - Roswell 

## Criteria for analysis 

Regional systems will be examined by aquifer unit. 
Each aquifer unit must have at least n=25 wells in both the pre-1910 and post-2010 time periods 


### Mississippi Embayment 

```{r}
# Pre-1910 clean file 

early1900s_me <- read_tsv(here::here("change_head", "data", "1900s_me.txt"))

# Modern 2010-2022 clean file 

modern_me <- read_tsv(here::here("change_head", "data", "modern_me.txt"))
```

##### Aquifer units to analyze in ME: 

- Middle Claiborne aquifer 
- Lower Claiborne confining unit 
- Middle Wilcox aquifer 

#### Middle Claiborne Aquifer

Post-2010 data 

```{r}
# Subtract median water level from topography to get the real water level 

modern_head_me <- modern_me %>% 
  mutate(head = topo_ft - median) 

# Keep only records in the mcaq, that are confined, and what we used in our regional analysis 

mca_mod <- modern_head_me %>% 
  filter(aquifer == "mcaq") %>% 
  filter(status == "confined") %>% 
  filter(!grepl("Unconfined single aquifer", code)) %>% 
  select(-vkbg_surf:-mdwy_surf) %>% 
  mutate(head_m = head*0.3048)

mca_mod_clean <- mca_mod %>% 
  select(site_no, head_m, artesian) %>% 
  mutate(time_period = "post-2010")  %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

mca_mod_clean$artesian <- as.character(mca_mod_clean$artesian)

```

Pre-1910 data

```{r}
# The DEM of the artesian flowing wells will be used as minimum head 
# Keep only records in the mcaq, that are confined, and what we used in our regional analysis

mca_1900s <- early1900s_me %>% 
  filter(aquifer == "mcaq") %>% 
  filter(status == "confined") %>% 
  mutate(non_art_head = topo_ft - abs(water_level_ft)) %>% 
  mutate(head = ifelse(artesian == 1, topo_ft, non_art_head)) %>% 
  select(-vkbg_surf_1:-mdwy_surf_1) %>% 
  mutate(head_m = head*0.3048)

mca_1900s_clean <- mca_1900s %>% 
  select(id, head_m, artesian) %>% 
  mutate(time_period = "pre-1910") %>% 
  rename(site_no = id) %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

mca_1900s_clean$artesian <- as.character(mca_1900s_clean$artesian)
```

Plotting 

```{r}

mca <- ggplot() + 
  geom_point(data = mca_1900s_clean, aes(x = x_axis, y = head_m, color = artesian), size = 2.5) +
  geom_line(data = mca_1900s_clean, aes(x = x_axis, y = head_m)) +
  geom_point(data = mca_mod_clean, aes(x = x_axis, y = head_m, color = artesian), shape = 15, size = 2.5) +
  geom_line(data = mca_mod_clean, aes(x = x_axis, y = head_m)) +
  theme_minimal() +
  labs(x = "", 
       y = "Hydraulic head (meters above sea level)
       ") +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(), 
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()) +
  scale_y_continuous(limits = c(-60, 150))

mca

ggsave(plot = mca, "mca.eps", path = "eps_files", width = 10, height = 5, units = "in")

```

Summary of data 

```{r}
# Final summaries 

# 1900s 
mca_1900_summary <- mca_1900s_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )

# Modern 

mca_mod_summary <- mca_mod_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )
```

```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(mca_1900_summary, caption = "Table 1. Mississippi Embayment, head of Middle Claiborne aquifer summary statistics for Pre-1910")

kable(mca_mod_summary, caption = "Table 1. Mississippi Embayment, head of Middle Claiborne aquifer summary statistics for Post-2010")

```

#### Lower Claiborne Confining Unit 

Post-2010 data 

```{r}
# Keep only records in the mcaq, that are confined, and what we used in our regional analysis 

lccu_mod <- modern_head_me %>% 
  filter(aquifer == "lccu") %>% 
  filter(status == "confined") %>% 
  filter(!grepl("Unconfined single aquifer", code)) %>% 
  select(-vkbg_surf:-mdwy_surf) %>% 
  mutate(head_m = head*0.3048)

lccu_mod_clean <- lccu_mod %>% 
  select(site_no, head_m, artesian) %>% 
  mutate(time_period = "post-2010")  %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

lccu_mod_clean$artesian <- as.character(lccu_mod_clean$artesian)

```

Pre-1910 data

```{r}
# The DEM of the artesian flowing wells will be used as minimum head 
# Keep only records in the mcaq, that are confined, and what we used in our regional analysis

lccu_1900s <- early1900s_me %>% 
  filter(aquifer == "lccu") %>% 
  filter(status == "confined") %>% 
  mutate(non_art_head = topo_ft - abs(water_level_ft)) %>% 
  mutate(head = ifelse(artesian == 1, topo_ft, non_art_head)) %>% 
  select(-vkbg_surf_1:-mdwy_surf_1) %>% 
  mutate(head_m = head*0.3048)

lccu_1900s_clean <- lccu_1900s %>% 
  select(id, head_m, artesian) %>% 
  mutate(time_period = "pre-1910") %>% 
  rename(site_no = id) %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

lccu_1900s_clean$artesian <- as.character(lccu_1900s_clean$artesian)

```

Plotting 

```{r}

lccu <- ggplot() + 
  geom_point(data = lccu_1900s_clean, aes(x = x_axis, y = head_m, color = artesian), size = 2.5) +
  geom_line(data = lccu_1900s_clean, aes(x = x_axis, y = head_m)) +
  geom_point(data = lccu_mod_clean, aes(x = x_axis, y = head_m, color = artesian), shape = 15, size = 2.5) +
  geom_line(data = lccu_mod_clean, aes(x = x_axis, y = head_m)) +
  theme_minimal() +
  labs(x = "", 
       y = "Hydraulic head (meters above sea level)
       ") +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(), 
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()) +
  scale_y_continuous(limits = c(-60, 150))

lccu

ggsave(plot = lccu, "lccu.eps", path = "eps_files", width = 10, height = 5, units = "in")

```

Summary of data 

```{r}
# Final summaries 

# 1900s 
lccu_1900_summary <- lccu_1900s_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )

# Modern 

lccu_mod_summary <- lccu_mod_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )
```

```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(lccu_1900_summary, caption = "Table 1. Mississippi Embayment, head of Lower Claiborne Confining unit summary statistics for Pre-1910")

kable(lccu_mod_summary, caption = "Table 1. Mississippi Embayment, head of Lower Claiborne Confining unit summary statistics for Post-2010")

```

#### Middle Wilcox Aquifer  

Post-2010 data 

```{r}
# Keep only records in the mcaq, that are confined, and what we used in our regional analysis 

mwaq_mod <- modern_head_me %>% 
  filter(aquifer == "mwaq") %>% 
  filter(status == "confined") %>% 
  filter(!grepl("Unconfined single aquifer", code)) %>% 
  select(-vkbg_surf:-mdwy_surf) %>% 
  mutate(head_m = head*0.3048)

mwaq_mod_clean <- mwaq_mod %>% 
  select(site_no, head_m, artesian) %>% 
  mutate(time_period = "post-2010")  %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

mwaq_mod_clean$artesian <- as.character(mwaq_mod_clean$artesian)

```

Pre-1910 data

```{r}
# The DEM of the artesian flowing wells will be used as minimum head 
# Keep only records in the mcaq, that are confined, and what we used in our regional analysis

mwaq_1900s <- early1900s_me %>% 
  filter(aquifer == "mwaq") %>% 
  filter(status == "confined") %>% 
  mutate(non_art_head = topo_ft - abs(water_level_ft)) %>% 
  mutate(head = ifelse(artesian == 1, topo_ft, non_art_head)) %>% 
  select(-vkbg_surf_1:-mdwy_surf_1) %>% 
  mutate(head_m = head*0.3048)

mwaq_1900s_clean <- mwaq_1900s %>% 
  select(id, head_m, artesian) %>% 
  mutate(time_period = "pre-1910") %>% 
  rename(site_no = id) %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

mwaq_1900s_clean$artesian <- as.character(mwaq_1900s_clean$artesian)

```

Plotting 

```{r}

mwaq <- ggplot() + 
  geom_point(data = mwaq_1900s_clean, aes(x = x_axis, y = head_m, color = artesian), size = 2.5) +
  geom_line(data = mwaq_1900s_clean, aes(x = x_axis, y = head_m)) +
  geom_point(data = mwaq_mod_clean, aes(x = x_axis, y = head_m, color = artesian), shape = 15, size = 2.5) +
  geom_line(data = mwaq_mod_clean, aes(x = x_axis, y = head_m)) +
  theme_minimal() +
  labs(x = "", 
       y = "Hydraulic head (meters above sea level)
       ") +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(), 
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()) +
  scale_y_continuous(limits = c(-60, 150))

mwaq

ggsave(plot = mwaq, "mwaq.eps", path = "eps_files", width = 10, height = 5, units = "in")

```

Summary of data 

```{r}
# Final summaries 

# 1900s 
mwaq_1900_summary <- mwaq_1900s_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )

# Modern 

mwaq_mod_summary <- mwaq_mod_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )
```

```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(mwaq_1900_summary, caption = "Table 1. Mississippi Embayment, head of Middle Wilcox aquifer summary statistics for Pre-1910")

kable(mwaq_mod_summary, caption = "Table 1. Mississippi Embayment, head of Middle Wilcox aquifer summary statistics for Post-2010")

```
