---
title: "Change in Head by Aquifer System"
author: "Annette Hilton"
date: '2023-06-18'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Attach Packages

```{r}
# Attach packages 

library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(knitr)


# Disable scientific notation 

options(scipen=999)
```

# Read in data 

- By each regional aquifer system that you can analyze based on criteria (n=25 samples)
- Data that has already been sorted in aquifer unit, contains DEM information 
- Regional aquifer systems include: 
  - Mississippi Embayment
  - North Atlantic Coastal Plain 
  - Central Valley 
  - Dakota Aquifer System
  - Floridan 
  - Houston-Gulf Coast 
  - Roswell
  - Columbia Plateau 

## Criteria for analysis 

Regional systems will be examined by aquifer unit. 
Each aquifer unit must have at least n=25 wells in both the pre-1910 and post-2010 time periods 


### Roswell Artesian Basin    

```{r}
# Pre-1910 clean file 

early1900s_ros <- read_tsv(here::here("change_head", "data", "1900s_roswell.txt"))

# Modern 2010-2022 clean file 

modern_ros <- read_tsv(here::here("change_head", "data", "modern_roswell.txt"))
```

##### Aquifer units to analyze in Roswell: 

- Artesia 

#### Artesia 

Post-2010 data 

```{r}
# Subtract median water level from topography to get the real water level 

roswell_mod <- modern_ros %>% 
  filter(aquifer == "artesia") %>% 
  mutate(head = topo_ft - median) %>% 
  mutate(head_m = head*0.3048) %>% 
  filter(!head > 6000) # Removed two outliers that we had previously agreed to remove from analysis (they are in the mountains, not Roswell basin) 

roswell_mod_clean <- roswell_mod %>% 
  select(site_no, head_m, artesian) %>% 
  mutate(time_period = "post-2010")  %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

roswell_mod_clean$artesian <- as.character(roswell_mod_clean$artesian)

```

Pre-1910 data

```{r}
# The DEM of the artesian flowing wells will be used as minimum head 

roswell_1900s <- early1900s_ros %>% 
  filter(aquifer == "artesia") %>% 
  mutate(non_art_head = topo_ft - abs(water_level_ft)) %>% 
  mutate(head = ifelse(artesian == 1, topo_ft, non_art_head)) %>%
  mutate(head_m = head*0.3048)

roswell_1900s_clean <- roswell_1900s %>% 
  select(id, head_m, artesian) %>% 
  mutate(time_period = "pre-1910") %>% 
  rename(site_no = id) %>% 
  arrange(desc(head_m)) %>% 
  mutate(row = row_number()) %>% 
  mutate(x_axis = row/n())

roswell_1900s_clean$artesian <- as.character(roswell_1900s_clean$artesian)
```

Plotting 

```{r}

roswell <- ggplot() + 
  geom_point(data = roswell_1900s_clean, aes(x = x_axis, y = head_m, color = artesian), size = 2.5) +
  geom_line(data = roswell_1900s_clean, aes(x = x_axis, y = head_m)) +
  geom_point(data = roswell_mod_clean, aes(x = x_axis, y = head_m, color = artesian), shape = 15, size = 2.5) +
  geom_line(data = roswell_mod_clean, aes(x = x_axis, y = head_m)) +
  theme_minimal() +
  theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank(), 
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()) 

roswell

ggsave(plot = roswell, "roswell.eps", path = "eps_files", width = 10, height = 5, units = "in")

```

Summary of data 

```{r}
# Final summaries 

# 1900s 
roswell_1900_summary <- roswell_1900s_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )

# Modern 

roswell_mod_summary <- roswell_mod_clean %>%  
  dplyr::summarise(
    mean = mean(head_m), 
    median = median(head_m),
    standard_deviation = sd(head_m)
    )
```

```{r echo = FALSE, results = 'asis'}

# Display knitted table 

kable(roswell_1900_summary, caption = "Table 1. Summary statistics for Pre-1910")

kable(roswell_mod_summary, caption = "Table 1. Summary statistics for Post-2010")

```

