---
title: "Test of Dakota Formation"
author: "Annette Hilton"
date: "02/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the Dakota Aquifer (South Dakota) as compared to present day (2010-2021).

The purpose of this Rmd is to: 

Test our calculation of the Dakota formation. 

Methods: 

1. We compiled lithological logs from the South Dakota Driller Database and selected well logs which referenced "Dakota" in them. 
2. Hand examined each lithological log for the top and bottom reported of the Dakota formation 
3. Other information was also compiled, such as the formation reported below the Dakota 
4. Plotted wells that reference the Dakota formation in ArcPro (lat/long)
5. Using NNIDW (Natural Neighbor Inverse Distance Weighted), created a raster layer of the Dakota formation 
  - Top layer (top of the Dakota) 
  - Bottom layer (bottom of the Dakota) 
6. Our layers of the Dakota formation are checked (below) against known wells to fall within Dakota formation from the SD Monitoring wells 
  
  

## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)

# Disable scientific notation 

options(scipen=999)
```

## Testing known Dakota wells with our raster layers of the Dakota formation 

### Read in Data from Rmd `sd_modern_data_cleaning.Rmd` 

South Dakota observation wells with aquifer notation (known wells screened in the Dakota formation; not used in our lithological log raster compilation) 

```{r}
sd_dakota_rastercheck <- readr::read_tsv(here::here("output_data", "sd", "sd_dakota_rastercheck.txt"))

```

Select only wells screened in Dakota aquifer 

```{r}
# Dakota aquifer code: "DKOT" 

sd_dakota <- sd_dakota_rastercheck %>% 
  select(id, lat, long, aquifer, depth) %>% 
  filter(aquifer == "DKOT")

write_tsv(sd_dakota, here::here("output_data", "sd", "sd_dakota_test.txt"))
```

### Process in ArcPRO 

*Note:* Both IDW and NNIDW were tested. NNIDW was ultimately selected for final analyses. 

1. "Extract multi values to points" tool to intersect the Dakota formation observation wells with DEM 
2. "Extract multi values to points" tool to intersect the Dakota formation wells with our versions of the top, top without Black Hills, and bottom layers of the Dakota formation 
3. Export table 

```{r}
# Read in processed file from ArcPro 

# IDW with entire top (including black hills), bottom, and top without black hills 

sd_dakota_rasterprocessed <- readr::read_csv(here::here("data", "south_dakota", "sd_wells_dakota_intersect.txt"))

#NNDW with top without black hills and bottom 

sd_dakota_natural <- readr::read_csv(here::here("data", "south_dakota", "test_dakota_natural.txt"))
```

```{r}
#IDW 
# True depth (elevation - well depth) 

test_depth <- sd_dakota_natural %>% 
  mutate(true_depth = topo_ft - depth)

# Between our layers of the Dakota formation? 

test_dakota_formation <- test_depth %>% 
  rowwise() %>% 
  mutate(dakota = ifelse(true_depth <= top_dak & true_depth >= bottom_dak, "full_dak", 
         ifelse(true_depth <= without_blackhills & true_depth >=bottom_dak, "full_dak_bh", 
                ifelse(true_depth<= without_blackhills, "top_only", NA)))) %>% 
    mutate(dakota_nat = ifelse(true_depth <= Natural_top_7 & true_depth >= Natural_bottom, "full_dak", 
                ifelse(true_depth<= Natural_top_7, "top_only", NA)))


# Summary 

summary <- test_dakota_formation %>% 
  group_by(dakota, dakota_nat) %>% 
  count(dakota, dakota_nat)
```


```{r}
# NATURAL NEIGHBOR 
# True depth (elevation - well depth) 

test_depth_nn <- sd_dakota_natural %>% 
  mutate(true_depth = topo_ft - depth)

# Between our layers of the Dakota formation? 

test_dakota_formation_nn <- test_depth_nn %>% 
  rowwise() %>% 
  mutate(dakota = ifelse(true_depth <= Natural_top_7 & true_depth >= Natural_bottom, "full_dak", 
                ifelse(true_depth < Natural_top_7, "top_only", 
                       ifelse(true_depth < Natural_top_7 & is.na(Natural_bottom), "top_nobottom", NA))))


test_dakota_form_case <- test_depth %>% 
  rowwise() %>% 
  mutate(dakota = case_when(true_depth <= top_dak & true_depth >= bottom_dak ~"full_dak",
                            true_depth <= without_blackhills & true_depth >= bottom_dak ~"full_dak_bh", 
                            true_depth <= without_blackhills ~"top_only"))

# Summary 

summary <- test_dakota_form_case %>% 
  group_by(dakota) %>% 
  count(dakota)

# Case when test 

test_dakota_form_nn_case <- test_depth_nn %>% 
  rowwise() %>% 
  mutate(dakota = case_when(true_depth <= Natural_top_7 & true_depth >= Natural_bottom ~"full_dak",
                            true_depth <= Natural_top_7 ~"top_only"))


# Summary 

summary_nn <- test_dakota_form_nn_case %>% 
  group_by(dakota) %>% 
  count(dakota)
```
