---
title: "South Dakota: Modern Well Records (~2010-2020)"
author: "Annette Hilton"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# South Dakota Modern Analysis 

This Rmd compiles modern groundwater data from South Dakota. 
South Dakota USGS data are sparse, necessary to pull in South Dakota drilling wells and observation monitoring wells from the South Dakota Department of Agriculture and Natural Resources. 

## Attach Packages

```{r}
# Attach packages 

library(parzer)
library(Rcpp)
library(tidyverse)
library(here)
library(janitor)
library(purrr)
library(broom)
library(tibble)
library(dplyr)
library(tidyr)
library(ggpubr)
library(data.table)
library(lubridate)


# Disable scientific notation 

options(scipen=999)
```

# Methods 

1. Read in modern data of South Dakota 
  a. Drilling wells from South Dakota Department of Agriculture and Natural Resources (https://apps.sd.gov/nr68welllogs/). From 1/01/1800-11/22/2021 
  b. Observation wells from South Dakota Department of Agriculture and Natural Resources (https://apps.sd.gov/nr69obswell/default.aspx#viewHelp). All records up to 11/18/2021
  c. USGS data for South Dakota 
2. Tidy data 
3. Prepare for Arc 

## Read in Data

### Drilling wells 

Pos = below land surface 
Neg = artesian 

```{r}
sd_raw_drill <-
    list.files(path = here::here("data", "south_dakota", "sd_drill_wells"),
               pattern = "*.csv", 
               full.names = T) %>% 
    map_df(~read_csv(., col_types = cols(.default = "c"))) 

```

## Tidy Data 

```{r}
# Tidy data 

sd_tidy_drill <- sd_raw_drill %>% 
  clean_names() %>% 
  select(-guid, -comp_rpt_pdf) %>% 
  mutate(date = lubridate::mdy(well_comp_date), 
         year = lubridate::year(date)) %>% 
  filter(year >= 2010) %>% 
  filter(!is.na(static_value), 
         !is.na(depth)) %>% 
  mutate(id = seq(1:2598), 
         artesian = ifelse(static_value <= 0, 1, 0)) 

sd_tidy_drill$id <- sub("^", "w", sd_tidy_drill$id) 

# Make consistent with other data frame 

sd_drill_join <- sd_tidy_drill %>% 
  rename(original_date = well_comp_date, 
         median_level = static_value) %>% 
  mutate(elevation = NA) %>% 
  select(id, lat, long, aquifer, elevation, original_date, date, year, depth, median_level, artesian) 

sd_drill_join$median_level <- as.numeric(sd_drill_join$median_level)

```

## Read in Data 

### Observation Wells From: https://apps.sd.gov/nr69obswell/default.aspx#viewHelp

Pos = artesian 
Neg = below land surface 

```{r}
# Well Readings--water level, dates 

sd_wells_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "well_readings"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))

# Metadata--well depth 

sd_depth_raw <- list.files(path = here::here("data", "south_dakota", "sd_obs_wells", "meta_data"), 
                                             pattern = "*.csv", 
                                             full.names = T) %>% 
                             map_df(~read_csv(., col_types = cols(.default = "c")))
```

## Tidy Data 

Join dataframes 

**CHANGE NEGATIVE VALUES TO POSITIVE TO BE CONSISTENT WITH DRILLING WELLS & USGS 
(Now, POS = below land surface, NEG = artesian)

```{r}
# Full join 

sd_join <- sd_wells_raw %>% 
  full_join(sd_depth_raw, by = "Observation Well")

write_tsv(sd_join, here::here("output_data", "sd_obs_wells.txt"))

# Tidy data 

sd_full_tidy <- sd_join %>% 
  clean_names() %>% 
  select(-is_manual, -x8, -x14) %>% 
  mutate(date = lubridate::mdy(reading_date), 
         year = year(date)) %>% 
  filter(year >= 2010) %>% 
  filter(!is.na(reading)) %>% 
  filter(!is.na(total_depth)) 

sd_full_tidy$reading <- as.numeric(sd_full_tidy$reading)

sd_full_med <- sd_full_tidy %>% 
  mutate(corrected_water_lev = (reading*-1)) %>% 
  group_by(observation_well) %>% 
  mutate(median_level = median(corrected_water_lev)) %>% 
  distinct(observation_well, .keep_all = TRUE) %>% 
  mutate(artesian = ifelse(median_level <= 0, 1, 0))


# Make consistent with other data frame 

sd_obs_join <- sd_full_med %>% 
  rename(original_date = reading_date, 
         lat = latitude, 
         long = longitude, 
         id = observation_well, 
         depth = total_depth) %>% 
  select(id, lat, long, aquifer, elevation, original_date, date, year, depth, median_level, artesian)

```


For data check of raster Dakota formation --only wells screened in Dakota aquifer 
01/31/22

```{r}
sd_dakota <- sd_obs_join %>% 
  select(id, lat, long, aquifer, depth) %>% 
  filter(aquifer == "DKOT")

write_tsv(sd_dakota, here::here("output_data", "sd", "sd_dakota_test.txt"))
```

## Read in Data 

### USGS 

Pos = below land surface 
Neg = artesian 

```{r}
sd_usgs <- read_tsv(here::here("output_data", "sd", "usgs_modern_sd.txt"))

```

```{r}
# Tidy data 

sd_tidy_usgs <- sd_usgs %>%   
  filter(level_year >= 2010) 

# Make consistent with other data frame 

sd_usgs_join <- sd_tidy_usgs %>% 
  rename(original_date = lev_dt, 
         date = level_date,
         year = level_year, 
         median_level = median, 
         lat = dec_lat_va, 
         long = dec_long_va, 
         id = site_no, 
         depth = well_depth_va) %>% 
  mutate(elevation = NA, 
         aquifer = NA) %>% 
  select(id, lat, long, aquifer, elevation, original_date, date, year, depth, median_level, artesian) 

sd_usgs_join$median_level <- as.numeric(sd_usgs_join$median_level)
sd_usgs_join$lat <- as.character(sd_usgs_join$lat)
sd_usgs_join$long <- as.character(sd_usgs_join$long)
sd_usgs_join$depth <- as.character(sd_usgs_join$depth)

```

Join both dataframes 

```{r}
sd_full <- rbind(sd_obs_join, sd_drill_join, sd_usgs_join)
```

Write for ArcGIS 

```{r}
write_tsv(sd_full, here::here("output_data", "sd", "sd_full_modern.txt"))
```


