---
title: "South Dakota Project--Well Classification"
author: "Annette Hilton"
date: "02/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the Dakota Aquifer (South Dakota) as compared to present day (2010-2021).

The purpose of this Rmd is to: 

1) Prep data for input to ArcGIS (for hydrogeological analysis & DEM): data prepped in `sd_modern_data_cleaning.Rmd` 
2) After ArcGIS processing: 

1) Determine well-aquifer classification 
2) Compare artesian conditions 
  

## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)

# Disable scientific notation 

options(scipen=999)
```

## ArcGIS Data Analysis 

The following steps were performed in ArcGIS: 

Hydrogeological data: 
Arcfile = "cv" 

1. Hydrogeologic layers (top and bottom) of Dakota aquifer (as created by us; lithological logs) 
2. Modern wells from SD (combination observation wells and USGS wells) (processed in Rmd `sd_modern_data_cleaning.Rmd`) and early 1900s data (early1900s_dem_wgs84_final.txt)
3. Use tool "Extract multi values to points" to determine well location intersection with geological raster layers
4. Export resulting well data as a text files

## Data Analysis (Post-Arc)

### Hydrogeological data 

```{r}
# Read in modern data file from Arc 

sd_mod_fromarc <- readr::read_csv(here::here("data", "south_dakota", "modern_dakota.txt"))

# Read in early 1900s data file from Arc (had to use read.csv rather than read_csv because default number of rows to check? Returned NAs rather than values) 

sd_early1900s_fromarc <- read.csv(here::here("data", "south_dakota", "early1900s_dakota_5.csv"))

# Read in meta file for early 1900s 

early1900s_meta <- readr::read_tsv(here::here("output_data", "early1900s_arc.txt"))

```

Tidy data 
**Note: ArcGIS somehow messes up data (water level, well depth)--from DEM, only keep the site ID and the DEM value. For 1900s data, join with old file here. 
Modern file already joined above. 

```{r}
# Basic data tidying 

# Modern data

sd_mod_tidy <- sd_mod_fromarc %>% 
  clean_names() %>%  
  select(-oid, -fid)

# 1900s data 

sd_early1900s_tidy <- sd_early1900s_fromarc %>% 
  clean_names() %>% 
  select(-oid1, -i_oid)

# Join with meta file 

sd_early1900s_join <- sd_early1900s_tidy %>% 
  full_join(early1900s_meta, by = "id") %>% 
  clean_names() %>% 
  select(-latitude_x:-artesian_x)

```


### SD Aquifer Information 

Hydrologic Model Layer Surfaces:

- Dakota (Top) 
- Dakota (Bottom) 


### Aquifer analysis 

- Find well bottom 

```{r}
# Convert `rastervalu` column from meters to feet (conversion: 1 meter = 3.28084 feet)
# Add a column that is elevation minus well depth
# Add a column that is elevation minus well water level 

# Modern data 

sd_analysis_mod <- sd_mod_tidy %>% 
  mutate(topo_ft = ned10m_bil * 3.28084) %>% 
  mutate(well_bottom = topo_ft - depth) %>% 
   mutate(keep = 
           ifelse(is.na(dakota_top_nnidw) & is.na(dakota_bottom_nnidw), "NA", "keep")) %>% 
  filter(!keep == "NA")

 
# 1900 data 

sd_analysis_1900 <- sd_early1900s_join %>% 
  mutate(topo_ft = ned10m_bilinear_cusa_albers102003 * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_ft_y) %>% 
  mutate(keep = 
           ifelse(is.na(dakota_top_nnidw) & is.na(dakota_bottom_nnidw), "NA", "keep")) %>% 
  filter(!keep == "NA")

```

#### Classify each well to an aquifer unit 

Theory: 

well bottom < top of Dakota = confined aquifer unit (in the Dakota) 


Top Only Analysis 

```{r}
# Classify wells as falling into the Dakota, or not 

# Modern data 

sd_class_mod <- sd_analysis_mod %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom <= dakota_top_nnidw, "dakota", NA))

# 1900 data 

sd_class_1900 <- sd_analysis_1900 %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom <= dakota_top_nnidw, "dakota", NA))

```

Sensitivity Analysis (Top and Bottom) 

```{r}
# Classify wells as falling into the Dakota, or not, based on top and bottom of Dakota raster layers 

# Modern data 

sd_class_mod_tb <- sd_analysis_mod %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom <= dakota_top_nnidw & well_bottom >= dakota_bottom_nnidw, "dakota", NA))

# 1900 data 

sd_class_1900_tb <- sd_analysis_1900 %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom <= dakota_top_nnidw & well_bottom >= dakota_bottom_nnidw, "dakota", NA))

```


---------------------------------------------------------------
Confined criteria coded out for now
---------------------------------------------------------------

<!-- #### Confined Well Criteria  -->

<!-- Confined aquifer unit:  -->

<!-- - Criteria: Must have at least 10 m (33 ft) of confining material above -->
<!-- - Code: If the aquifer is confined, calculate the sum of the confining unit thickness above (layer4 & layer5). If the sum of the confining unit thickness is >= 33 ft (10m), then "keep" the well.  -->


<!-- First, determine thickness of aquifer units  -->

<!-- ```{r} -->
<!-- # Thickness of corcoran clay -->

<!-- # Modern data  -->

<!-- cv_thickness_mod <- cv_class_mod %>%  -->
<!--   mutate( -->
<!--     corcoran_thick = layer4_top - layer6_top) -->

<!-- # 1900 data  -->

<!-- cv_thickness_1900 <- cv_class_1900 %>%  -->
<!--   mutate( -->
<!--     corcoran_thick = layer4_top - layer6_top) -->

<!-- ``` -->

<!-- Then, determine cumulative thickness & which wells meet criteria  -->

<!-- ```{r} -->
<!-- # Modern data  -->

<!-- cv_cum_thickness_mod <- cv_thickness_mod %>%  -->
<!--   rowwise() %>%  -->
<!--   dplyr::mutate(keep =  -->
<!--            ifelse(corcoran_thick >= 33, "keep", NA)) -->

<!-- # Modern data  -->

<!-- cv_cum_thickness_1900 <- cv_thickness_1900 %>%  -->
<!--   rowwise() %>%  -->
<!--   dplyr::mutate(keep =  -->
<!--            ifelse(corcoran_thick >= 33, "keep", NA)) -->
<!-- ``` -->


#### Comparison of artesian conditions 

Dakota Top Only 

Dakota 1900 to 2010-2020 

Modern n = 64
1900s n = 175

```{r}
# Modern 
modern_sd <- sd_class_mod %>% 
  filter(aquifer == "dakota") 

# Join with file that separates drilling wells out (only keeps monitoring wells and USGS wells)
# Read in file (from `sd_modern_data_cleaning.Rmd`)
sd_usgsmon_mod <- readr::read_tsv(here::here("output_data", "sd", "sd_usgsmon_modern.txt"))

# Join files 
modern_sd_joined <- sd_usgsmon_mod %>% 
  inner_join(modern_sd, by = "id", all = FALSE)

# write file for arc 

write_tsv(modern_sd_joined, here::here("output_data", "sd", "modern_sd_limited.txt"))

# 1900s 
historical_sd <- sd_class_1900 %>% 
  filter(aquifer == "dakota")

# write file for arc 

write_tsv(historical_sd, here::here("output_data", "sd", "historical_sd.txt"))

```


Dakota Top and Bottom 

Dakota 1900 to 2010-2020 

Modern n = 31
1900s n = 64

```{r}
# Modern 
modern_sd_tb <- sd_class_mod_tb %>% 
  filter(aquifer == "dakota") 

# Join files 
modern_sd_joined_tb <- sd_usgsmon_mod %>% 
  inner_join(modern_sd_tb, by = "id", all = FALSE)

# write file for arc 

write_tsv(modern_sd_joined_tb, here::here("output_data", "sd", "modern_sd_tb_limited.txt"))

# 1900s 
historical_sd_tb <- sd_class_1900_tb %>% 
  filter(aquifer == "dakota")

# write file for arc 

write_tsv(historical_sd_tb, here::here("output_data", "sd", "historical_sd_tb.txt"))

```
