---
title: "Mississippi Embayment Project--Well Classification"
author: "Annette Hilton"
date: "12/07/2021"
output: html_docu/ment
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the Mississippi Embayment (ME) as compared to present day (2010-2021).



## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)


# Disable scientific notation 

options(scipen=999)
```

```{r}
# Modern data 
# Read in modern USGS data w/ DEM 
dem_usgs_modern <- readr::read_csv(here::here("data", "dem", "dem_usgs_modern.csv")) 

# Read in modern USGS data full file (from Rmd `modern_data_cleaning.Rmd`)
usgs_modern <- readr::read_tsv(here::here("output_data", "usgs_modern.txt"))


# 1900s data 
# Read in early 1900s data w/ DEM 
dem_1900s <- readr::read_csv(here::here("data", "dem", "dem_1900s.csv"))

# Read in early 1900s data full file (from Rmd `early_1900s_cleaning.Rmd`)
early_1900s <- readr::read_tsv(here::here("output_data", "early1900s_arc.txt"))
```

### Join DEM files with full data files 

```{r}
# Join modern 

modern_join <- usgs_modern %>% 
  full_join(dem_usgs_modern, by = "site_no")

# Join 1900s 

early1900s_join <- early_1900s %>% 
  full_join(dem_1900s, by = "id")
```

## Data Analysis (Pre-Arc)

### Refine dataset for the Mississippi Embayment 

```{r}
# Only keep well data for states in ME: 
# Missouri, Illinois, Kentucky, Tennessee, Arkansas, Mississippi, Alabama, Louisiana

me <- modern_join %>% 
  filter(id.x %in% c("al", "ar", "il", "ky", "la", "ms", "mo", "tn"))

```


### Prep data for import into ArcGIS 

```{r}
# Write to text file to then export to ArcGIS

write_tsv(me, here::here("output_data", "me", "me_modern_toclassify.txt"))

# Write early 1900s file w/ DEM to export into ArcGIS 

write_tsv(early1900s_join, here::here("output_data", "me", "early1900s_toclassify.txt"))
  
```


## ArcGIS Data Analysis 

The following steps were performed in ArcGIS: 

Hydrogeological data: 
Arcfile = "miss_embayment" 

1. Import USGS ME geologic data (rasters) 
2. Import USGS wells (text file from analysis above; 'unique_me_wells.txt') 
3. Use tool "Extract multi values to points" to determine well location intersection with geological raster layers 
4. Export data as text file  


## Data Analysis (Post-Arc)

### Hydrogeological data 

```{r}
# Read in modern data file from Arc 

me_mod_fromarc <- readr::read_csv(here::here("data", "me", "archived", "me_modern_classifyreturn.txt"))

# Read in early 1900s data file from Arc 

early1900s_fromarc <- readr::read_csv(here::here("data", "me", "archived", "early1900s_classifyreturn.txt"))

```

Tidy data 

```{r}
# Basic data tidying 
# Create `na_strings` to indicate pesky NA values
# Replace NA numeric indicators with "NA" 

# Modern data

na_strings <- c("-9999", "-9999.00", "-9999.000", "-9999.0000", "-9999.00000", "-9999.000000")

me_mod_tidy <- me_mod_fromarc %>% 
  clean_names() %>% 
  select(-oid) %>% 
  replace_with_na_all(condition = ~.x %in% na_strings)

# 1900s data 

early1900s_tidy <- early1900s_fromarc %>% 
  clean_names() %>% 
  select(-oid)

```

### ME Aquifer Information 

Hydrologic Model Layer Surfaces:

- Mississippi river valley aquifer **mrva** 
- Vicksburg-Jackson Group (top) **(vkbg)**   
- Upper Claiborne aquifer (top) **(ucaq)**
- Middle Claiborne confining unit (top) **(mccu)**  
- Middle Claiborne aquifer (top) **(mcaq)** 
- Lower Claiborne confining unit (top) **(lccu)**  
- Lower Claiborne aquifer (top) **(lcaq)** 
- Middle Wilcox aquifer (top) **(mwaq)**
- Lower Wilcox aquifer (top) **(lwaq)**
- Midway confining unit (top) **(mdwy)** 

Unit Range (ft): 

- mrva: unknown 
- vkbg: 566 (-364) 
- ucaq: 521 (-874)
- mccu: 540 (-1461)
- mcaq: 654 (-1659)
- lccu: 614 (-2539)
- lcaq: 616 (-3102)
- mwaq: 632 (-3347)
- lwaq: 555 (-2320)
- mdwy: 553 (-6036)


### Aquifer analysis 

- Find well bottom 
- Remove observations that do not fall within ME (all NAs) 

```{r}
# Convert `rastervalu` column from meters to feet (conversion: 1 meter = 3.28084 feet)
# Add a column that is elevation minus well depth
# Add column to remove rows if vkbg:mdwy are all NAs (does not intersect with ME) 

# Modern data 

me_analysis_mod <- me_mod_tidy %>% 
  mutate(topo_ft = rastervalu * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth) %>% 
  mutate(keep = 
           ifelse(is.na(vkbg_surf) & is.na(ucaq_surf) & is.na(mccu_surf) 
                  & is.na(mcaq_surf) & is.na(lccu_surf) & is.na(lcaq_surf)
                  & is.na(mwaq_surf) & is.na(lwaq_surf) & is.na(mdwy_surf), "NA", "keep")) %>% 
  filter(!keep == "NA")

# Write to tsv file for analysis in other Rmds 

# write_tsv(me_analysis, here::here("me_data", "me_analysis.txt"))


# 1900s data 

me_analysis_1900 <- early1900s_tidy %>% 
  mutate(topo_ft = rastervalu * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_ft) %>% 
  mutate(keep = 
           ifelse(is.na(vkbg_surf) & is.na(ucaq_surf) & is.na(mccu_surf) 
                  & is.na(mcaq_surf) & is.na(lccu_surf) & is.na(lcaq_surf)
                  & is.na(mwaq_surf) & is.na(lwaq_surf) & is.na(mdwy_surf), "NA", "keep")) %>% 
  filter(!keep == "NA")

# Write to tsv file for analysis in other Rmds 

# write_tsv(me_analysis, here::here("me_data", "me_analysis.txt"))
  
```

#### Classify each well to an aquifer unit 


Process: 

If well bottom is greater than second unit (top), then it falls within the first unit (second unit top is the bottom of the first unit). 

Theory: 

well bottom > vkbg (first unit) = mrva 
well bottom > ucaq (second unit) = vkbg (first unit) 
well bottom > mccu (third unit) = ucaq (second unit) 
well bottom > mcaq (fourth unit) = mccu (third unit) 
well bottom > lccu (fifth unit) = mcaq (fourth unit) 
well bottom > lcaq (sixth unit) = lccu (fifth unit) 
well bottom > mwaq (seventh unit) = lcaq (sixth unit) 
well bottom > lwaq (eigth unit) = mwaq (seventh unit) 
well bottom > mdwy (ninth unit) = lwaq (eigth unit) 
well bottom < mdwy (ninth unit) = mdwy (ninth unit) 

Code: 

well bottom < topo_ft & well bottom > max(vkbg:mdwy) = mrva
well bottom < vkbg & well bottom > max(ucaq:mdwy) = vkbg
well bottom < ucaq & well bottom > max(mccu:mdwy) = ucaq
well bottom < mccu & well bottom > max(mcaq:mdwy) = mccu 
well bottom < mcaq & well bottom > max(lccu:mdwy) = mcaq
well bottom < lccu & well bottom > max(lcaq:mdwy) = lccu 
well bottom < lcaq & well bottom > max(mwaq:mdwy) = lcaq
well bottom < mwaq & well bottom > max(lwaq:mdwy) = mwaq
well bottom < lwaq & well bottom > mdwy = lwaq
well bottom < mdwy = mdwy 


```{r}
# Code for assigning each well to an aquifer unit 
# Not elegant but currently works! 

# Modern data 

me_class_mod <- me_analysis_mod %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom < topo_ft & well_bottom > max(vkbg_surf, ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf, lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mrva", 
        ifelse(well_bottom < vkbg_surf & well_bottom > max(ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE),"vkbg",
        ifelse(well_bottom < ucaq_surf & well_bottom > max(mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "ucaq", 
        ifelse(well_bottom < mccu_surf & well_bottom > max(mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mccu",
        ifelse(well_bottom < mcaq_surf & well_bottom > max(lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mcaq", 
        ifelse(well_bottom < lccu_surf & well_bottom > max(lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lccu", 
        ifelse(well_bottom < lcaq_surf & well_bottom > max(mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lcaq",
        ifelse(well_bottom < mwaq_surf & well_bottom > max(lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mwaq",
        ifelse(well_bottom < lwaq_surf & well_bottom > mdwy_surf, "lwaq",
        ifelse(well_bottom < mdwy_surf, "mdwy", NA)))))))))))


# 1900s data 

me_class_1900 <- me_analysis_1900 %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom < topo_ft & well_bottom > max(vkbg_surf, ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf, lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mrva", 
        ifelse(well_bottom < vkbg_surf & well_bottom > max(ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE),"vkbg",
        ifelse(well_bottom < ucaq_surf & well_bottom > max(mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "ucaq", 
        ifelse(well_bottom < mccu_surf & well_bottom > max(mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mccu",
        ifelse(well_bottom < mcaq_surf & well_bottom > max(lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mcaq", 
        ifelse(well_bottom < lccu_surf & well_bottom > max(lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lccu", 
        ifelse(well_bottom < lcaq_surf & well_bottom > max(mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lcaq",
        ifelse(well_bottom < mwaq_surf & well_bottom > max(lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mwaq",
        ifelse(well_bottom < lwaq_surf & well_bottom > mdwy_surf, "lwaq",
        ifelse(well_bottom < mdwy_surf, "mdwy", NA)))))))))))

```

Summary of wells per aquifer unit 

```{r}
# Summary of wells per aquifer unit 

# Modern data 
summarymod_me_old <- me_class_mod %>% 
  group_by(aquifer) %>% 
  count(aquifer) 

write_tsv(summary_me, here::here("me_data", "summary_me.txt"))


# 1900s data 

summary1900_me_old <- me_class_1900 %>% 
  group_by(aquifer) %>% 
  count(aquifer)

# join summaries 

full_summary <- summary1900_me %>% 
  full_join(summarymod_me, by = "aquifer") %>% 
  rename(historical = n.x, 
         modern = n.y)

```

#### Add criteria to include wells only drilled in confining unit by 10 feet 

For aquifer MCAQ (wells in confining unit LCCU) 


```{r}
# Modern data (adds 7 wells)
confining_depth_mod <- me_class_mod %>% 
  mutate(testing = 
           ifelse(aquifer == "lccu" & well_bottom > (lccu_surf - 10), "c.mcaq", NA))

# 1900s data (adds 1 well)
confining_depth_1900<- me_class_1900 %>% 
  mutate(testing = 
           ifelse(aquifer == "lccu" & well_bottom > (lccu_surf - 10), "c.mcaq", NA))
```

#### Confined Well Criteria 

Confined aquifer units (UCAQ, MCAQ, LCAQ, MWAQ, LWAQ): 

- Criteria: Must have at least 10 m (33 ft) of confining material above
- Code: Dependent on the aquifer, calculate the sum of the confining unit thickness above. If the sum of the confining unit thickness is >= 33 ft (10m), then "keep" the well. 


First, determine thickness of aquifer units 

```{r}
# Thickness of aquifer units 

# Modern data 

me_thickness_mod <- me_class_mod %>% 
  mutate(
    vkbg_thick = vkbg_surf - ucaq_surf, 
    ucaq_thick = ucaq_surf - mccu_surf, 
    mccu_thick = mccu_surf - mcaq_surf, 
    mcaq_thick = mcaq_surf - lccu_surf, 
    lccu_thick = lccu_surf - lcaq_surf, 
    lcaq_thick = lcaq_surf - mwaq_surf, 
    mwaq_thick = mwaq_surf - lwaq_surf, 
    lwaq_thick = lwaq_surf - mdwy_surf
  )

# 1900s data 

me_thickness_1900 <- me_class_1900 %>% 
  mutate(
    vkbg_thick = vkbg_surf - ucaq_surf, 
    ucaq_thick = ucaq_surf - mccu_surf, 
    mccu_thick = mccu_surf - mcaq_surf, 
    mcaq_thick = mcaq_surf - lccu_surf, 
    lccu_thick = lccu_surf - lcaq_surf, 
    lcaq_thick = lcaq_surf - mwaq_surf, 
    mwaq_thick = mwaq_surf - lwaq_surf, 
    lwaq_thick = lwaq_surf - mdwy_surf
  )

```

Then, determine cumulative thickness & which wells meet criteria 

```{r}
# Only consider confining units for thickness 
# VKBG, MCCU, LCCU

# Modern 

me_cum_thickness_mod <- me_thickness_mod %>% 
  rowwise() %>% 
    dplyr::mutate(cum_thick = 
        ifelse(aquifer == "mrva", sum(vkbg_thick, mccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "ucaq", sum(vkbg_thick), 
        ifelse(aquifer == "mcaq", sum(vkbg_thick, mccu_thick, na.rm = TRUE),
        ifelse(aquifer == "lcaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "mwaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "lwaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
               NA))))))) %>% 
  dplyr::mutate(keep = 
           ifelse(cum_thick >= 33, "keep", NA))


# 1900s 

me_cum_thickness_1900 <- me_thickness_1900 %>% 
  rowwise() %>% 
    dplyr::mutate(cum_thick = 
        ifelse(aquifer == "mrva", sum(vkbg_thick, mccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "ucaq", sum(vkbg_thick), 
        ifelse(aquifer == "mcaq", sum(vkbg_thick, mccu_thick, na.rm = TRUE),
        ifelse(aquifer == "lcaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "mwaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
        ifelse(aquifer == "lwaq", sum(vkbg_thick, mccu_thick, lccu_thick, na.rm = TRUE), 
               NA))))))) %>% 
  dplyr::mutate(keep = 
           ifelse(cum_thick >= 33, "keep", NA))
```


### Separate each aquifer/unit and confined 

Confined aquifers 
* Does not include confining unit 

Modern n = 205
1900s n = 675

```{r}
# All aquifers that are confined by our criteria  

# Modern data 
me_confined_mod <- me_cum_thickness_mod %>% 
  filter(keep == "keep" & !aquifer == "mrva") %>% 
  mutate(artesian = 
           ifelse(lev_va < 0, 1, 0))

# Write as tsv 
write_tsv(me_confined_mod, here::here("output_data", "me", "me_confined_mod.txt"))

# 1900 data 
me_confined_1900 <- me_cum_thickness_1900 %>% 
  filter(keep == "keep" & !aquifer == "mrva")

# Write as tsv 
write_tsv(me_confined_1900, here::here("output_data", "me", "me_confined_historical.txt"))

```


#### Comparison of artesian conditions 

MCAQ (Middle Claiborne Aquifer) 1900 to 2010-2020 

Modern = 779 
1900 = 149 

```{r}
# Separate MCAQ & C.MCAQ units 

# Modern 
modern_mcaq <- confining_depth_mod %>% 
  filter(aquifer == "mcaq" | testing == "c.mcaq") %>% 
  mutate(artesian = 
           ifelse(lev_va < 0, 1, 0))

# write file for arc 

write_tsv(modern_mcaq, here::here("output_data", "me", "modern_mcaq.txt"))

# 1900s 
historical_mcaq <- confining_depth_1900 %>% 
  filter(aquifer == "mcaq" | testing == "c.mcaq") 

# write file for arc 

write_tsv(historical_mcaq, here::here("output_data", "me", "historical_mcaq.txt"))

```


MCAQ (Middle Claiborne Aquifer) & LCCU 1900 to 2010-2020 

Modern n = 849
1900 n = 250

```{r}
# Separate MCAQ & LCCU units 

# Modern 
modern_mcaq_lccu <- me_cum_thickness_mod %>% 
  filter(aquifer == "mcaq" | aquifer == "lccu") %>% 
  mutate(artesian = 
           ifelse(lev_va < 0, 1, 0))

# write file for arc 

write_tsv(modern_mcaq_lccu, here::here("output_data", "me", "modern_mcaq_lccu.txt"))

# 1900s 
historical_mcaq_lccu <- me_cum_thickness_1900 %>% 
  filter(aquifer == "mcaq" | aquifer == "lccu") 

# write file for arc 

write_tsv(historical_mcaq_lccu, here::here("output_data", "me", "historical_mcaq_lccu.txt"))

```



------------------------------------------------------------------
```{r}
# Graph these differences 

# join summaries 

full_summary <- summary1900_me %>% 
  full_join(summarymod_me, by = "aquifer") %>% 
  rename(historical = n.x, 
         modern = n.y)

write_tsv(full_summary, "full_summary_aquifers.txt")

# 
# 
# # Graph 
# 
# ggplot(full_summary) + 
#   geom_point(aes(x = aquifer, y = historical)) + 
#   geom_segment(aes(x = aquifer, xend = aquifer, y = 0, yend = historical)) +
#   geom_point(aes(x = aquifer, y = modern)) +
#   geom_segment(aes(x = aquifer, xend = aquifer, y = 0, yend = modern))
# 
# ggplot(data = full_summary) + 
#   geom_bar(aes(x = aquifer, y = historical), position = position_dodge())
# 
# 
# 
# 
# ggplot(data = full_summary, aes(x = aquifer, y = historical, fill = modern)) +
#   geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75)
```

# Introduction

This project is an examination of early 1900s (~1890-1906) groundwater conditions in the Mississippi Embayment (ME) as compared to present day (2010-2021).

The purpose of this Rmd is to: 

1) Determine well-aquifer classification 
2) Compare artesian conditions 
  

## Attach Packages and Data 

```{r}
# Attach packages 

library(plyr) # Be sure to load `plyr()` first, otherwise other functions in tidyverse break! 
library(tidyverse)
library(here)
library(janitor)
library(naniar)

# Disable scientific notation 

options(scipen=999)
```

## ArcGIS Data Analysis 

The following steps were performed in ArcGIS: 

Hydrogeological data: 
Arcfile = "miss_embayment_classify" (ArcMap) 

1. Hydrogeologic layers (raster files) 
2. Modern wells from ME (USGS) (processed in Rmd `modern_data_cleaning.Rmd`) and early 1900s data (early1900s_dem_wgs84_final.txt)
3. Use tool "Extract multi values to points" to determine well location intersection with geological raster layers
4. Export resulting well data as a text files

## Data Analysis (Post-Arc)

### Hydrogeological data 

```{r}
# Read in modern data file from Arc 

me_mod_fromarc_1 <- readr::read_csv(here::here("data", "me", "me_mod_classify.txt")) %>% 
  clean_names()

# Read in early 1900s data file from Arc 

early1900s_fromarc <- readr::read_csv(here::here("data", "me", "me_historical_classify.txt"), guess_max = 6653) %>% 
  clean_names()

```

### ME Aquifer Information 

Hydrogeologic Model Layer Surfaces:

- Mississippi river valley aquifer **mrva** 
- Vicksburg-Jackson Group (top) **(vkbg)**   
- Upper Claiborne aquifer (top) **(ucaq)**
- Middle Claiborne confining unit (top) **(mccu)**  
- Middle Claiborne aquifer (top) **(mcaq)** 
- Lower Claiborne confining unit (top) **(lccu)**  
- Lower Claiborne aquifer (top) **(lcaq)** 
- Middle Wilcox aquifer (top) **(mwaq)**
- Lower Wilcox aquifer (top) **(lwaq)**
- Midway confining unit (top) **(mdwy)** 

Unit Range (ft): 

- mrva: unknown 
- vkbg: 566 (-364) 
- ucaq: 521 (-874)
- mccu: 540 (-1461)
- mcaq: 654 (-1659)
- lccu: 614 (-2539)
- lcaq: 616 (-3102)
- mwaq: 632 (-3347)
- lwaq: 555 (-2320)
- mdwy: 553 (-6036)


### Aquifer analysis 

- Find well bottom 

```{r}
# Convert `rastervalu` column from meters to feet (conversion: 1 meter = 3.28084 feet)
# Add a column that is elevation minus well depth
# Add a column that is elevation minus well water level 

# Modern data 

me_analysis_mod_1 <- me_mod_fromarc_1 %>% 
  mutate(topo_ft = rastervalu * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_va) %>% 
  mutate(keep = 
           ifelse(is.na(vkbg_surf) & is.na(ucaq_surf) & is.na(mccu_surf) 
                  & is.na(mcaq_surf) & is.na(lccu_surf) & is.na(lcaq_surf)
                  & is.na(mwaq_surf) & is.na(lwaq_surf) & is.na(mdwy_surf), "NA", "keep")) %>% 
  filter(!keep == "NA")

 
# 1900 data 

me_analysis_1900 <- early1900s_fromarc %>% 
  mutate(topo_ft = ned10m_bilinear_cusa_albers102003 * 3.28084) %>% 
  mutate(well_bottom = topo_ft - well_depth_ft) %>% 
  mutate(keep = 
           ifelse(is.na(vkbg_surf_1) & is.na(ucaq_surf_1) & is.na(mccu_surf_1) 
                  & is.na(mcaq_surf_1) & is.na(lccu_surf_1) & is.na(lcaq_surf_1)
                  & is.na(mwaq_surf_1) & is.na(lwaq_surf_1) & is.na(mdwy_surf_1), "NA", "keep")) %>% 
  filter(!keep == "NA")

```

#### Classify each well to an aquifer unit 


Process: 

If well bottom is greater than second unit (top), then it falls within the first unit (second unit top is the bottom of the first unit). 

Theory: 

well bottom > vkbg (first unit) = mrva 
well bottom > ucaq (second unit) = vkbg (first unit) 
well bottom > mccu (third unit) = ucaq (second unit) 
well bottom > mcaq (fourth unit) = mccu (third unit) 
well bottom > lccu (fifth unit) = mcaq (fourth unit) 
well bottom > lcaq (sixth unit) = lccu (fifth unit) 
well bottom > mwaq (seventh unit) = lcaq (sixth unit) 
well bottom > lwaq (eigth unit) = mwaq (seventh unit) 
well bottom > mdwy (ninth unit) = lwaq (eigth unit) 
well bottom < mdwy (ninth unit) = mdwy (ninth unit) 

Code: 

well bottom < topo_ft & well bottom > max(vkbg:mdwy) = mrva
well bottom < vkbg & well bottom > max(ucaq:mdwy) = vkbg
well bottom < ucaq & well bottom > max(mccu:mdwy) = ucaq
well bottom < mccu & well bottom > max(mcaq:mdwy) = mccu 
well bottom < mcaq & well bottom > max(lccu:mdwy) = mcaq
well bottom < lccu & well bottom > max(lcaq:mdwy) = lccu 
well bottom < lcaq & well bottom > max(mwaq:mdwy) = lcaq
well bottom < mwaq & well bottom > max(lwaq:mdwy) = mwaq
well bottom < lwaq & well bottom > mdwy = lwaq
well bottom < mdwy = mdwy 


```{r}
# Code for assigning each well to an aquifer unit 
# Not elegant but currently works! 

# Modern data 

me_class_mod_1 <- me_analysis_mod_1 %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom < topo_ft & well_bottom > max(vkbg_surf, ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf, lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mrva", 
        ifelse(well_bottom < vkbg_surf & well_bottom > max(ucaq_surf, mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE),"vkbg",
        ifelse(well_bottom < ucaq_surf & well_bottom > max(mccu_surf, 
                                                            mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "ucaq", 
        ifelse(well_bottom < mccu_surf & well_bottom > max(mcaq_surf, lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mccu",
        ifelse(well_bottom < mcaq_surf & well_bottom > max(lccu_surf,lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mcaq", 
        ifelse(well_bottom < lccu_surf & well_bottom > max(lcaq_surf,
                                                            mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lccu", 
        ifelse(well_bottom < lcaq_surf & well_bottom > max(mwaq_surf, lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "lcaq",
        ifelse(well_bottom < mwaq_surf & well_bottom > max(lwaq_surf, mdwy_surf, 
                                                            na.rm = TRUE), "mwaq",
        ifelse(well_bottom < lwaq_surf & well_bottom > mdwy_surf, "lwaq",
        ifelse(well_bottom < mdwy_surf, "mdwy", NA)))))))))))


# 1900s data 

me_class_1900 <- me_analysis_1900 %>% 
  rowwise() %>%
  mutate(aquifer = 
        ifelse(well_bottom < topo_ft & well_bottom > max(vkbg_surf_1, ucaq_surf_1, mccu_surf_1, 
                                                            mcaq_surf_1, lccu_surf_1, lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE), "mrva", 
        ifelse(well_bottom < vkbg_surf_1 & well_bottom > max(ucaq_surf_1, mccu_surf_1, 
                                                            mcaq_surf_1, lccu_surf_1, lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE),"vkbg",
        ifelse(well_bottom < ucaq_surf_1 & well_bottom > max(mccu_surf_1, 
                                                            mcaq_surf_1, lccu_surf_1, lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE), "ucaq", 
        ifelse(well_bottom < mccu_surf_1 & well_bottom > max(mcaq_surf_1, lccu_surf_1, lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1,
                                                            na.rm = TRUE), "mccu",
        ifelse(well_bottom < mcaq_surf_1 & well_bottom > max(lccu_surf_1, lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE), "mcaq", 
        ifelse(well_bottom < lccu_surf_1 & well_bottom > max(lcaq_surf_1,
                                                            mwaq_surf_1, lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE), "lccu", 
        ifelse(well_bottom < lcaq_surf_1 & well_bottom > max(mwaq_surf_1, lwaq_surf_1, mdwy_surf_1,
                                                            na.rm = TRUE), "lcaq",
        ifelse(well_bottom < mwaq_surf_1 & well_bottom > max(lwaq_surf_1, mdwy_surf_1, 
                                                            na.rm = TRUE), "mwaq",
        ifelse(well_bottom < lwaq_surf_1 & well_bottom > mdwy_surf_1, "lwaq",
        ifelse(well_bottom < mdwy_surf_1, "mdwy", NA)))))))))))

```

Summary of wells per aquifer unit 

```{r}
# Summary of wells per aquifer unit 

# Modern data 
summarymod_me_1 <- me_class_mod_1 %>% 
  ungroup() %>% 
  group_by(aquifer) %>% 
  count(aquifer) 


# 1900s data 

summary1900_me <- me_class_1900 %>% 
  group_by(aquifer) %>% 
  count(aquifer)

```



**STOP HERE** 4/21/22



### Separate each aquifer/unit and confined 

Confined aquifers 
* Does not include confining unit 

Modern n = 205
1900s n = 675

```{r}
# All aquifers that are confined by our criteria  

# Modern data 
me_confined_mod <- me_cum_thickness_mod %>% 
  filter(keep == "keep" & !aquifer == "mrva") %>% 
  mutate(artesian = 
           ifelse(lev_va < 0, 1, 0))

# Write as tsv 
write_tsv(me_confined_mod, here::here("output_data", "me", "me_confined_mod.txt"))

# 1900 data 
me_confined_1900 <- me_cum_thickness_1900 %>% 
  filter(keep == "keep" & !aquifer == "mrva")

# Write as tsv 
write_tsv(me_confined_1900, here::here("output_data", "me", "me_confined_historical.txt"))

```


#### Comparison of artesian conditions 

1900 to 2010-2020 

Modern n = 143
1900s n = 386

```{r}
# MODERN 

# Modern all 
modern_roswell_all <- roswell_class_mod %>% 
  filter(!is.na(aquifer))

# write file for arc 
write_tsv(modern_roswell_all, here::here("output_data", "roswell", "modern_roswell_all.txt"))

# Modern alluvium 
modern_roswell_alluvium <- modern_roswell_all %>% 
  filter(aquifer == "alluvium") 

# write file for arc 
write_tsv(modern_roswell_alluvium, here::here("output_data", "roswell", "modern_roswell_alluvium.txt"))

# Modern artesia 
modern_roswell_artesia <- modern_roswell_all %>% 
  filter(aquifer == "artesia") 

# write file for arc 
write_tsv(modern_roswell_artesia, here::here("output_data", "roswell", "modern_roswell_artesia.txt"))


# 1900s 

# 1900s all 
historical_roswell_all <- roswell_class_1900 %>% 
  filter(!is.na(aquifer))

# write file for arc 
write_tsv(historical_roswell_all, here::here("output_data", "roswell", "historical_roswell_all.txt"))

# Historical alluvium 
historical_roswell_alluvium <- historical_roswell_all %>% 
  filter(aquifer == "alluvium") 

# write file for arc 
write_tsv(historical_roswell_alluvium, here::here("output_data", "roswell", "historical_roswell_alluvium.txt"))

# Historical artesia 
historical_roswell_artesia <- historical_roswell_all %>% 
  filter(aquifer == "artesia") 

# write file for arc 
write_tsv(historical_roswell_artesia, here::here("output_data", "roswell", "historical_roswell_artesia.txt"))
```



